---
title: "School-to-Work Pathways: Building Your Analysis File"
subtitle: "Module 0 — Preparing State Longitudinal Data for Sequence Analysis"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 5.5
)
```

---

# Overview

This is a preparatory module for researchers who are working with their own
state longitudinal data and want to replicate or adapt the sequence analysis
approach used in the Sequences to Success in Washington State project.

**If you are using the toy dataset provided in this repository, you can skip
this module.** The toy dataset is already structured and ready to use with
Modules 1, 2, and 3.

This module covers:

1. The structure your raw data needs to be in before you begin
2. Key decisions you will need to make, with the approach we took and why
3. Code for transforming raw enrollment and wage records into a
   sequence-ready wide format file
4. Alternative approaches for researchers whose data or research questions
   differ from ours
   
   
   
**Output files**
This module produces the following output files, which are saved to your output 
folder and are required for Modules 1, 2, and 3:

- combined_seq_long.csv — the combined enrollment and employment states in long format
- combined_seq_wide.csv — the combined states pivoted to wide format, used as the 
primary input for sequence analysis


**A note on scope:** This module is intentionally focused.
Rather than trying to cover every possible data structure or research design,
we walk through the specific decisions we made for the Sequences to Success
project and explain the reasoning behind each one. Where our approach may not
fit your context, we flag alternatives for you to explore.

**A note on package versions:** The code in this module was developed and tested 
using specific versions of R and its packages. If you encounter unexpected 
errors or differences in output, it may be due to package version differences. 
See the [Session information] section at the end of this module for the full 
list of package versions used in development. You can check your own package 
versions by running sessionInfo() in your RStudio console, and update packages 
by running update.packages().


---

# 1. Setup

## 1.1 Setting up your working directory

Before running any code in this module, you need to make sure R knows where to 
find your files. The working directory is the folder on your computer that R 
looks in by default when reading or saving files.

If you are using an R Project (recommended): Open the .Rproj file in the 
repository folder. RStudio will automatically set your working directory to the 
project folder, and all file paths in this module will work without any changes.

If you are not using an R Project: Set your working directory manually by 
running the following in your RStudio console, replacing the path with the 
location of your repository folder on your computer:

```{r set-working-directory}

setwd("path/to/your/repository/folder")

```

# For example, on a Mac: setwd("/Users/yourname/Documents/sequences-to-success")
# For example, on a PC: setwd("C:/Users/yourname/Documents/sequences-to-success")

You can also set the working directory by going to Session > Set Working 
Directory > Choose Directory in the RStudio menu and navigating to the repository 
folder.

## 1.2 Load packages

```{r load-packages}
library(tidyverse)   # dplyr, tidyr, purrr, etc.
library(scales)      # for formatting
library(knitr)       # for clean table output
library(kableExtra)  # for enhanced table formatting

# If any of these are not installed, run:
# install.packages(c("tidyverse", "scales", "knitr", "kableExtra"))
```

---

# 2. What your raw data needs to look like

Sequence analysis requires that you can place each student in a defined
**state** at each point in time. To do this, you need at minimum:

- An **enrollment file** with one row per student per enrollment spell (or
  per term), showing institution type and enrollment dates or term identifiers
- A **wage or employment file** with one row per student per quarter showing
  whether the student was employed and their earnings

Both files should share a common student identifier so they can be linked.

The toy dataset in this repository was generated to mimic these two source
files. If you are working with real state data, your enrollment file likely
comes from a postsecondary enrollment system (such as a state higher education
data system) and your wage file comes from Unemployment Insurance (UI) wage
records.

---

# 3. Key decisions

Before writing any code, you will need to make several decisions about how to
handle your data. This section describes the decisions we made for the
Sequences to Success project and the reasoning behind each one.

## 3.1 Defining the time spine

**The decision:** We used calendar quarters as our unit of time, with one
observation per student per quarter across a 6.5-year follow-up window (26
quarters total).

**Why we made this choice:** UI wage records are reported quarterly, so
quarters are the natural unit for employment data. Aligning enrollment to
the same quarterly spine made it straightforward to combine enrollment and
employment into a single state for each time period.

**How we handled term-to-quarter alignment:** Postsecondary enrollment is
reported by term (fall, winter, spring, summer) rather than by quarter. We
mapped each term to the calendar quarters it overlapped with. For example,
if spring term ran from February through May, a student enrolled in spring
term would be counted as enrolled in both Q1 (January-March) and Q2
(April-June). This means a student can be counted as enrolled in more
quarters than the number of terms they attended, which slightly overcounts
enrollment at term boundaries. We considered this an acceptable tradeoff for
the simplicity it provides.

**Alternatives to consider:**

- **Semester-based systems:** If your state uses semesters rather than
  quarters, you could define your time spine as semesters instead. You
  would then need to decide how to align wage data, which is quarterly,
  to a semester spine (e.g., assign wages to the semester with the most
  overlap, or average across the two quarters that fall within each semester).
- **Fall/Spring only:** Some researchers exclude summer and winter terms if
  enrollment in those terms is sparse or inconsistently reported. If you
  take this approach, note that students who attend primarily in summer or
  winter terms may be misclassified.
- **Shorter or longer follow-up windows:** Our analysis was limited to 6.5 years 
  due to data availability at the time. This was sufficient to capture 
  bachelor's degree completion for most students in our cohort. Depending on 
  your research question, a shorter window (e.g., 4 years) or longer window 
  may be more appropriate.


## 3.2 Handling concurrent enrollment

**The decision:** When a student was enrolled at both a community/technical
college (CTC) and a 4-year institution in the same quarter, we coded that
quarter as 4-year enrollment (4YR).

**Why we made this choice:** Concurrent enrollment may reflect a student
who is primarily pursuing a bachelor's degree while taking a course or two
at a CTC. Coding these students as 4YR preserves their likely credential
trajectory and avoids overstating CTC enrollment.

**Alternatives to consider:**

- **Code as CTC:** If your research question focuses on CTC participation,
  you may prefer to count any CTC enrollment, including concurrent
  enrollment, as CTC.
- **Create a separate "concurrent" state:** If concurrent enrollment is
  common in your data and you want to study it explicitly, you could add a
  seventh state (e.g., `Both`) to capture students enrolled at both types
  of institutions simultaneously.
- **Use the primary institution:** Some states designate a primary
  institution for students with concurrent enrollment. If that information
  is available, you could use it to assign a single institution type.

## 3.3 Combining enrollment and employment into a single state

**The decision:** For the toy data set, we combined enrollment status and 
employment status into a single state for each quarter, producing six possible 
states:

| State        | Meaning                                              |
|--------------|------------------------------------------------------|
| `4YR_Work`   | Enrolled at 4-year college AND employed              |
| `4YR_NoWork` | Enrolled at 4-year college, not employed             |
| `CTC_Work`   | Enrolled at CTC AND employed                         |
| `CTC_NoWork` | Enrolled at CTC, not employed                        |
| `Work_Only`  | Not enrolled, employed                               |
| `Neither`    | Not enrolled, not employed                           |

**Why we made this choice:** Combining enrollment and employment into a
single state captures a wide range of student experiences, including the
common pattern of working while enrolled. A student who attends a 4-year
college while working part-time has a meaningfully different experience from
one who attends full-time without working, and the combined state reflects
that difference. 

**NOTE:** For the larger project, we explored several versions of state
definitions, including concurrent work and enrollment. HOwever, our team
was ultimately focused on postsecondary pathways, so we landed on a version
of the states that prioritized enrollment and also captured credential 
attainment at each quarter (e.g., "enrolled in 4-YR with Associate's degree
already earned" or "working, not enrolled with bachelor's degree already 
earned").

**Alternatives to consider:**

- **Enrollment only:** If your research question focuses on postsecondary
  pathways rather than labor market connection, you could use enrollment
  status alone (CTC, 4YR, None) as your state. This produces a simpler
  sequence with fewer states and may be easier to interpret. Like our team
  did for our larger project, you could choose to also code credential 
  attainment into your states. 
- **Enrollment priority:** Similar to our approach for the project, you could
  use an enrollment-priority approach, where a student's state is
  determined by their enrollment status alone regardless of employment. 
  Employment can then be analyzed separately as an outcome. This is a cleaner 
  approach if you want to keep the pathway and outcome analyses separate.


## 3.4 Handling missing data

**The decision:** When a student did not appear in either the enrollment or
the wage records for a given quarter, we coded that quarter as `Neither`
(not enrolled, not employed).

**Why we made this choice:** Administrative records capture students when
they are active in a system. Absence from both records is the best available
evidence that a student was neither enrolled nor employed in that period.
Creating a separate "missing" state would conflate true absence from both
systems with data quality issues, and would complicate the sequence analysis
by adding a state that is difficult to interpret substantively.

**Alternatives to consider:**

- **Treat as missing:** If you have reason to believe that data gaps reflect
  true data quality issues rather than absence from both systems, you may
  want to flag these quarters as missing rather than coding them as Neither.
  TraMineR can handle missing values in sequences, though they require
  careful handling in distance computation.
- **Imputation:** For short gaps (e.g., a single quarter surrounded by
  enrollment quarters), you could impute the state based on surrounding
  quarters. This is most appropriate when gaps likely reflect administrative
  reporting lags rather than true stop-outs.

---

# 4. Transforming your data

This section provides code for transforming raw enrollment and wage records
into the combined state sequence file used in Modules 2 and 3. The code
is written to work with the toy dataset but is structured to be adaptable
to real state data.

## 4.1 Define the time spine

```{r define-spine}
# Define the quarterly time spine for the analysis.
# Adjust the start year, end year, and cohort graduation year to match your data.

cohort_grad_year <- 2016   # high school graduation year for your cohort
start_year       <- 2017   # first year of follow-up
end_year         <- 2023   # last year of follow-up

# Generate all quarters in the follow-up window
time_spine <- expand_grid(
  year    = start_year:end_year,
  quarter = 1:4
) %>%
  mutate(
    calendar_quarter = paste0(year, "Q", quarter),
    quarter_number   = row_number(),
    # Approximate start and end months for each quarter
    month_start = (quarter - 1) * 3 + 1,
    month_end   = quarter * 3
  )

# How many quarters are we tracking?
n_quarters <- nrow(time_spine)
cat("Tracking", n_quarters, "quarters from", start_year, "Q1 to", end_year, "Q4\n")
```
**Note:** We ultimately started our sequence data at the 3rd quarter of the 
graduation year to reflect that many students graduate in May or June and would
be unlikely to start attending college before Q3.

## 4.2 Map enrollment terms to quarters

This step converts term-based enrollment records into quarterly enrollment
indicators. Each term is mapped to every quarter it overlaps with.

```{r map-terms-to-quarters}
# Load the raw enrollment data
# Expected columns: student_id, term_start_month, term_start_year,
#                   term_end_month, term_end_year, institution_type (CTC or 4YR)
#
# For the toy dataset, we use the pre-built enrollment_long file which already
# has quarterly enrollment. If you are working with real data, you will need
# to map your term dates to quarters using the logic below.

data_dir <- "longitudinal_toy_data"
enrollment_long <- read_csv(file.path(data_dir, "enrollment_long.csv"))

# --- For real state data, use this approach: ---
#
# enrollment_quarterly <- enrollment_raw %>%
#   # For each enrollment record, identify which quarters it overlaps with
#   rowwise() %>%
#   mutate(quarters_enrolled = list(
#     time_spine %>%
#       filter(
#         # Quarter overlaps with term if term starts before quarter ends
#         # and term ends after quarter starts
#         !(term_end_year < year | (term_end_year == year & term_end_month < month_start)) &
#         !(term_start_year > year | (term_start_year == year & term_start_month > month_end))
#       ) %>%
#       pull(calendar_quarter)
#   )) %>%
#   unnest(quarters_enrolled) %>%
#   # Handle concurrent enrollment: if a student has both CTC and 4YR in
#   # the same quarter, keep only 4YR (our project decision — see Section 3.2)
#   group_by(student_id, quarters_enrolled) %>%
#   summarise(
#     institution_type = if_else(any(institution_type == "4YR"), "4YR", "CTC"),
#     .groups = "drop"
#   )
```

## 4.3 Create the quarterly employment indicator

```{r create-employment}
# Load the raw wage records
# Expected columns: student_id, calendar_quarter, quarterly_wages
#
# A student is considered employed in a quarter if they have positive wages
# in that quarter in the UI wage records.

wages_long <- read_csv(file.path(data_dir, "wage_records_long.csv"))

# Create a simple employed/not-employed indicator
# Adjust the minimum wage threshold to match your state's definition
MIN_QUARTERLY_WAGES <- 100  # minimum quarterly wages to be considered employed

employment_quarterly <- wages_long %>%
  mutate(
    employed = case_when(
      is.na(quarterly_wages)          ~ 0L,
      quarterly_wages >= MIN_QUARTERLY_WAGES ~ 1L,
      TRUE                            ~ 0L
    )
  ) %>%
  select(student_id, quarter_number, calendar_quarter, employed, quarterly_wages)

cat("Employment records:", nrow(employment_quarterly), "rows\n")
cat("Students with any employment:", 
    employment_quarterly %>% 
      filter(employed == 1) %>% 
      pull(student_id) %>% 
      n_distinct(), "\n")
```

## 4.4 Build the combined state sequence

This is the core transformation step. For each student and each quarter, we
combine their enrollment status and employment status into a single state.

```{r build-combined-states}
# Create a full student x quarter grid so every student has a row for
# every quarter, even if they do not appear in either source file
students <- read_csv(file.path(data_dir, "students.csv"))

full_grid <- expand_grid(
  student_id     = students$student_id,
  quarter_number = 1:n_quarters
) %>%
  left_join(
    time_spine %>% select(quarter_number, calendar_quarter),
    by = "quarter_number"
  )

# Join enrollment status onto the full grid
enroll_quarterly <- enrollment_long %>%
  select(student_id, term_number, enrollment_status) %>%
  rename(quarter_number = term_number)

full_grid <- full_grid %>%
  left_join(enroll_quarterly, by = c("student_id", "quarter_number")) %>%
  # Students not in the enrollment file in a given quarter are not enrolled
  mutate(enrollment_status = replace_na(enrollment_status, "None"))

# Join employment status onto the full grid
full_grid <- full_grid %>%
  left_join(
    employment_quarterly %>% select(student_id, quarter_number, employed),
    by = c("student_id", "quarter_number")
  ) %>%
  # Students not in the wage file in a given quarter are not employed
  # (see Section 3.4 for discussion of this decision)
  mutate(employed = replace_na(employed, 0L))

# Assign the combined state
combined_seq_long <- full_grid %>%
  mutate(
    combined_state = case_when(
      enrollment_status == "4YR" & employed == 1 ~ "4YR_Work",
      enrollment_status == "4YR" & employed == 0 ~ "4YR_NoWork",
      enrollment_status == "CTC" & employed == 1 ~ "CTC_Work",
      enrollment_status == "CTC" & employed == 0 ~ "CTC_NoWork",
      enrollment_status == "None" & employed == 1 ~ "Work_Only",
      enrollment_status == "None" & employed == 0 ~ "Neither"
    )
  ) %>%
  select(student_id, quarter_number, calendar_quarter, combined_state)

# Quick check: state distribution
combined_seq_long %>%
  count(combined_state) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  arrange(desc(n)) %>%
  kable(caption = "Combined state distribution across all students and quarters") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 4.5 Pivot to wide format

Modules 2 and 3 require the sequence data in wide format, with one row per
student and one column per quarter.

```{r pivot-wide}
combined_seq_wide <- combined_seq_long %>%
  select(student_id, quarter_number, combined_state) %>%
  pivot_wider(
    names_from  = quarter_number,
    names_prefix = "cstate_",
    values_from = combined_state
  )

cat("Wide format dimensions:", nrow(combined_seq_wide), "rows x",
    ncol(combined_seq_wide), "columns\n")

# Preview first few rows and columns
combined_seq_wide[1:5, 1:8]
```

## 4.6 Save the output

```{r save-output}
# Save the combined sequence files for use in Modules 2 and 3.
# If you are working with real state data, adjust the output path as needed.

output_dir <- "longitudinal_toy_data"

#write_csv(combined_seq_long, file.path(output_dir, "combined_seq_long.csv"))
#write_csv(combined_seq_wide, file.path(output_dir, "combined_seq_wide.csv"))

cat("Files saved to", output_dir, "\n")
```

---

# 5. Checks before moving on

Before proceeding to Module 1, it is worth running a few quick checks to
make sure your analysis file looks as expected.

```{r checks}
# 1. Do all students have the expected number of quarters?
combined_seq_long %>%
  count(student_id) %>%
  summarise(
    min_quarters  = min(n),
    max_quarters  = max(n),
    all_complete  = all(n == n_quarters)
  ) %>%
  kable(caption = "Completeness check: quarters per student") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

# 2. Are there any unexpected state values?
combined_seq_long %>%
  filter(!combined_state %in% c("4YR_Work", "4YR_NoWork", "CTC_Work",
                                 "CTC_NoWork", "Work_Only", "Neither")) %>%
  nrow() %>%
  { cat("Rows with unexpected state values:", ., "\n") }

# 3. What share of students are ever enrolled? Ever employed?
combined_seq_long %>%
  group_by(student_id) %>%
  summarise(
    ever_enrolled = any(combined_state %in% c("4YR_Work", "4YR_NoWork",
                                               "CTC_Work", "CTC_NoWork")),
    ever_employed = any(combined_state %in% c("4YR_Work", "CTC_Work",
                                               "Work_Only"))
  ) %>%
  summarise(
    pct_ever_enrolled = percent(mean(ever_enrolled), accuracy = 0.1),
    pct_ever_employed = percent(mean(ever_employed), accuracy = 0.1)
  ) %>%
  kable(caption = "Share of students ever enrolled and ever employed") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

# 4. What does employment look like in the final year?
combined_seq_long %>%
  filter(quarter_number %in% (n_quarters - 3):n_quarters) %>%
  group_by(quarter_number) %>%
  summarise(
    pct_employed = percent(
      mean(combined_state %in% c("4YR_Work", "CTC_Work", "Work_Only")),
      accuracy = 0.1
    )
  ) %>%
  kable(caption = "Employment rate in final four quarters") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

---

# Next steps

Once you have confirmed that your combined sequence file looks reasonable,
you are ready to move on to the core modules:

- **Module 1** (`01_data_prep_and_visualization.Rmd`) — Descriptive summaries
  of demographics, enrollment patterns, credential attainment, and labor
  market outcomes
- **Module 2** (`02_sequence_analysis.Rmd`) — Defining and visualizing
  sequences, computing pairwise distances using Optimal Matching
- **Module 3** (`03_cluster_analysis.Rmd`) — Clustering sequences into
  pathway groups and examining outcomes by cluster
  
  
**You have now built your analysis file.** The following files have been saved 
and are ready to use in the analysis modules:

- combined_seq_long.csv — used for descriptive summaries in Module 1
- combined_seq_wide.csv — used as the primary input for sequence analysis in Module 2

Open 01_data_prep_and_visualization.Rmd to continue to the next step, where you 
will explore the structure and patterns of your data before moving into the 
sequence analysis.

---

# Session information

The code in this module was developed and tested using the package versions 
listed below. If you encounter unexpected errors, checking your package 
versions against these is a good first troubleshooting step. You can update 
packages by running `update.packages()` in your RStudio console.

```{r session-info}
sessionInfo()
```
**Development Session Info**
R version 4.5.2 (2025-10-31)
Platform: x86_64-pc-linux-gnu
Running under: Red Hat Enterprise Linux 9.8 (Plow)

Matrix products: default
BLAS/LAPACK: FlexiBLAS OPENBLAS-OPENMP;  LAPACK version 3.9.0

locale:
LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8   
LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C           LC_TELEPHONE=C        
LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

time zone: America/New_York
tzcode source: system (glibc)

attached base packages:
stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
kableExtra_1.4.0 knitr_1.50       scales_1.4.0     lubridate_1.9.4  forcats_1.0.1    stringr_1.6.0    dplyr_1.1.4     
purrr_1.2.0      readr_2.1.6      tidyr_1.3.1      tibble_3.3.0     ggplot2_4.0.1    tidyverse_2.0.0 

loaded via a namespace (and not attached):
sass_0.4.10        generics_0.1.4     xml2_1.4.1         stringi_1.8.7      hms_1.1.4          digest_0.6.38     
magrittr_2.0.4     evaluate_1.0.5     grid_4.5.2         timechange_0.3.0   RColorBrewer_1.1-3 fastmap_1.2.0     
jsonlite_2.0.0     viridisLite_0.4.2  jquerylib_0.1.4    textshaping_1.0.4  cli_3.6.5          rlang_1.1.6       
crayon_1.5.3       bit64_4.6.0-1      cachem_1.1.0       withr_3.0.2        yaml_2.3.10        tools_4.5.2       
parallel_4.5.2     tzdb_0.5.0         vctrs_0.6.5        R6_2.6.1           lifecycle_1.0.4    bit_4.6.0         
vroom_1.6.6        pkgconfig_2.0.3    bslib_0.9.0        pillar_1.11.1      gtable_0.3.6       rsconnect_1.6.1   
glue_1.8.0         systemfonts_1.3.1  xfun_0.54          tidyselect_1.2.1   rstudioapi_0.17.1  farver_2.1.2      
htmltools_0.5.8.1  rmarkdown_2.30     svglite_2.2.2      compiler_4.5.2     S7_0.2.1          

