---
title: "School-to-Work Pathways: Data Preparation and Visualization"
subtitle: "Module 1 of 3 — Using Synthetic Longitudinal Data"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,     # show all code by default
  message = FALSE,    # suppress package load messages
  #warning = FALSE,    # suppress warnings in output
  fig.width  = 9,
  fig.height = 5.5
)
```

---

# Overview

This is the first of three modules in a series on using **sequence analysis** and 
**cluster analysis** to study school-to-work pathways in longitudinal administrative data.

This module covers:

1. Loading and inspecting the synthetic longitudinal dataset
2. Understanding the structure of the raw data tables
3. Exploring demographic and high school characteristics
4. Describing postsecondary enrollment and credential patterns
5. Summarizing labor market outcomes by pathway


**Output files**
This module does not produce any output files. It is focused on exploring and 
understanding the data before moving into the analysis. All files needed for 
Modules 2 and 3 are either provided in the longitudinal_toy_data/ folder or 
produced by Module 0.


**Data note:** All data used in these modules are fully synthetic and were 
generated for teaching purposes only. The patterns are illustrative and do not 
represent any real state, institution, or individual.


**A note on package versions:** The code in this module was developed and tested 
using specific versions of R and its packages. If you encounter unexpected 
errors or differences in output, it may be due to package version differences. 
See the [Session information] section at the end of this module for the full 
list of package versions used in development. You can check your own package 
versions by running sessionInfo() in your RStudio console, and update packages 
by running update.packages().

---

# 1. Setup

## 1.1 Load packages

```{r load-packages}
# Core data manipulation and visualization
library(tidyverse)   # dplyr, ggplot2, tidyr, purrr, etc.
library(scales)      # for formatting axes (dollar signs, percents, etc.)
library(knitr)       # for clean table display with kable()
library(kableExtra)  # for enhanced table formatting

# If any of these are not installed, run:
# install.packages(c("tidyverse", "scales", "knitr", "kableExtra"))
```

## 1.2 Load the data

The synthetic dataset consists of several linked tables. We load each one 
separately and then join them as needed for analysis. All tables share a 
common `student_id` key.

```{r load-data}
# Adjust this path to match the location of the folder where you have stored the data files.
data_dir <- "/data/share/cdi/Sequences to Success in WA/Quantitative/GitHub Repo/longitudinal_toy_data"

# --- Core tables ---

# One row per student; static demographic characteristics
students <- read_csv(file.path(data_dir, "students.csv"))

# One row per student; high school GPA, FRPL, diploma type, dual credit flag
hs_records <- read_csv(file.path(data_dir, "hs_records.csv"))

# One row per student × term; enrollment status each term (CTC, 4YR, None)
enrollment_long <- read_csv(file.path(data_dir, "enrollment_long.csv"))

# One row per credential earned; certificate, associate's, or bachelor's
credentials <- read_csv(file.path(data_dir, "credentials.csv"))

# One row per student × quarter; quarterly wages and employment indicator
wages_long <- read_csv(file.path(data_dir, "wage_records_long.csv"))

# wide version
wages_wide <- read_csv(file.path(data_dir, "wage_records_wide.csv"))


# --- Pre-built analysis-ready table ---

# One row per student with demographics, HS record, credential summary,
# and aggregate labor market outcomes merged together
analysis_flat <- read_csv(file.path(data_dir, "analysis_flat.csv"))

# --- Sequence-ready wide tables (used in Modules 2 and 3) ---

# One column per term showing enrollment state (wide format for TraMineR)
enrollment_wide <- read_csv(file.path(data_dir, "enrollment_wide.csv"))

# One column per term showing combined enrollment + employment state
combined_wide <- read_csv(file.path(data_dir, "combined_seq_wide.csv"))

#long version
combined_long <- read_csv(file.path(data_dir, "combined_seq_long.csv"))
```

---

# 2. Data structure

Before analyzing, it is important to understand the shape and content of each 
table. This section inspects each one briefly.

## 2.1 Students table

```{r inspect-students}
# glimpse() is the tidyverse equivalent of str() — it shows column names,
# types, and the first few values in a compact format
glimpse(students)
```

```{r students-n}
# How many students are in the cohort?
nrow(students)
```

## 2.2 Enrollment table (long format)

The enrollment table is in **long format**: each student has one row per term. 
This is the natural format for administrative records and is easy to reshape 
for visualization or sequence analysis.

```{r inspect-enrollment}
glimpse(enrollment_long)

# Preview: one student's enrollment history across all terms
enrollment_long %>%
  filter(student_id == "STU000001") %>%
  select(term_label, enrollment_status)
```

## 2.3 Wage records (long format)

```{r inspect-wages}
glimpse(wages_long)

# Preview: one student's wage history across all quarters
wages_long %>%
  filter(student_id == "STU000016") %>%
  select(calendar_quarter, employed, quarterly_wages)
```

## 2.4 Analysis-ready flat file

The flat file merges everything into one row per student. This is convenient 
for regression models, cross-tabulations, and outcome comparisons — but it 
collapses the time dimension, so it cannot be used for sequence analysis.

```{r inspect-flat}
glimpse(analysis_flat)
```

---

# 3. Demographic characteristics

## 3.1 Race and ethnicity

```{r race-eth-table}
students %>%
  count(race_ethnicity) %>%
  mutate(
    pct = n / sum(n),
    pct_label = percent(pct, accuracy = 0.1)
  ) %>%
  arrange(desc(n)) %>%
  kable(
    col.names = c("Race/Ethnicity", "N", "Proportion", "Percent"),
    caption   = "Distribution of race/ethnicity in the synthetic cohort"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r race-eth-plot}
students %>%
  count(race_ethnicity) %>%
  mutate(
    race_ethnicity = fct_reorder(race_ethnicity, n),
    pct = n / sum(n)
  ) %>%
  ggplot(aes(x = race_ethnicity, y = pct)) +
  geom_col(fill = "#2E86AB", alpha = 0.85) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.6)) +
  labs(
    title = "Race/Ethnicity Distribution",
    x     = NULL,
    y     = "Share of cohort"
  ) +
  theme_minimal(base_size = 13)
```

## 3.2 Key demographic indicators

```{r demo-summary}
# Summarize binary demographic indicators as percentages
students %>%
  summarise(
    `% Female`      = mean(sex == "Female"),
    `% FRPL`        = mean(frpl == 1),
    `% Rural`       = mean(rural_flag == 1),
    `% First-gen`   = mean(first_gen == 1),
    `% IEP`         = mean(iep_flag == 1),
    `% ELL`         = mean(ell_flag == 1)
  ) %>%
  pivot_longer(everything(), names_to = "Indicator", values_to = "Proportion") %>%
  mutate(Percent = percent(Proportion, accuracy = 0.1)) %>%
  kable(caption = "Key demographic indicators") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

# 4. High school characteristics

## 4.1 GPA distribution

```{r gpa-dist}
hs_records %>%
  ggplot(aes(x = hs_gpa)) +
  geom_histogram(binwidth = 0.1, fill = "lightblue", color = "white", alpha = 0.85) +
  geom_vline(aes(xintercept = median(hs_gpa, na.rm = TRUE)),
             color = "#E84855", linetype = "dashed", linewidth = 1) +
  annotate("text", x = median(hs_records$hs_gpa, na.rm = TRUE) + 0.05,
           y = 120, label = paste0("Median: ",
             round(median(hs_records$hs_gpa, na.rm = TRUE), 2)),
           color = "darkred", hjust = 0, size = 4) +
  labs(
    title = "High School GPA Distribution",
    x     = "High school GPA",
    y     = "Number of students"
  ) +
  theme_minimal(base_size = 13)
```

## 4.2 Dual credit participation

```{r dual-credit}
hs_records %>%
  summarise(
    `Students with dual credit` = sum(dual_credit_flag == 1, na.rm = TRUE),
    `Total students`            = n(),
    `% with dual credit`        = percent(mean(dual_credit_flag == 1, na.rm = TRUE),
                                           accuracy = 0.1)
  ) %>%
  kable(caption = "Dual credit participation") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

---

# 5. Postsecondary enrollment patterns

## 5.1 Ever enrolled in postsecondary education

```{r ever-enrolled}
# A student "ever enrolled" if they have at least one term with CTC or 4YR status
ever_enrolled <- enrollment_long %>%
  group_by(student_id) %>%
  summarise(
    ever_enrolled = any(enrollment_status %in% c("CTC", "4YR"), na.rm = TRUE)
  )

ever_enrolled %>%
  count(ever_enrolled) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  kable(col.names = c("Ever enrolled in PSE?", "N", "Percent"),
        caption   = "Postsecondary enrollment rate") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 5.2 Enrollment state distribution over time

This chart shows the share of students in each enrollment state at each term. 
It is sometimes called a **state distribution plot** or **tempogram**. You will 
see a more polished version of this in Module 2 using TraMineR, but it is useful 
to build it manually first to understand what is being plotted.

```{r enrollment-over-time}
# Step 1: count students in each state at each term
# We treat NA as "Unknown" for this chart
state_dist <- enrollment_long %>%
  mutate(enrollment_status = replace_na(enrollment_status, "Unknown")) %>%
  count(term_label, enrollment_status) %>%
  group_by(term_label) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup()

# Step 2: order terms chronologically using the term_number from the long table
term_order <- enrollment_long %>%
  distinct(term_number, term_label) %>%
  arrange(term_number)

state_dist <- state_dist %>%
  mutate(term_label = factor(term_label, levels = term_order$term_label))

# Step 3: plot
state_dist %>%
  filter(enrollment_status != "Unknown") %>%
  ggplot(aes(x = term_label, y = pct, fill = enrollment_status)) +
  geom_col() +
  scale_fill_manual(
    values = c("4YR" = "#2E86AB", "CTC" = "#A23B72", "None" = "#E0E0E0"),
    labels = c("4YR" = "4-Year College", "CTC" = "Community/Technical College",
               "None" = "Not Enrolled")
  ) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title = "Enrollment State Distribution by Term",
    x     = "Term",
    y     = "Share of cohort",
    fill  = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

## 5.3 Credential attainment

```{r credential-summary}
credentials %>%
  count(credential_type) %>%
  mutate(pct = percent(n / nrow(students), accuracy = 0.1)) %>%
  kable(col.names = c("Credential type", "N students", "% of cohort"),
        caption   = "Credential attainment by type") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

---

# 6. Labor market outcomes

## 6.1 Overall employment rates

```{r employment-overall}
# "Stably employed" is defined as employed all 4 quarters of the final study year
# with total annual wages >= $14,000 (mirrors definition used for our project)
analysis_flat %>%
  summarise(
    `Ever employed`    = percent(mean(ever_employed,    na.rm = TRUE), accuracy = 0.1),
    `Stably employed`  = percent(mean(stably_employed,  na.rm = TRUE), accuracy = 0.1),
    `Median annual wages (stably employed)` =
      dollar(median(annual_wages_2023[stably_employed == 1], na.rm = TRUE))
  ) %>%
  kable(caption = "Overall labor market outcomes") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## 6.2 Wage distribution (stably employed)

```{r wage-dist}
analysis_flat %>%
  filter(stably_employed == 1) %>%
  ggplot(aes(x = annual_wages_2023)) +
  geom_histogram(binwidth = 3000, fill = "#2E86AB", color = "white", alpha = 0.85) +
  geom_vline(aes(xintercept = median(annual_wages_2023, na.rm = TRUE)),
             color = "#E84855", linetype = "dashed", linewidth = 1) +
  scale_x_continuous(labels = dollar_format(scale = 1/1000, suffix = "K")) +
  labs(
    title    = "Annual Wage Distribution (Stably Employed Students)",
    subtitle = "Dashed line = median",
    x        = "Annual wages",
    y        = "Number of students"
  ) +
  theme_minimal(base_size = 13)
```

## 6.3 Employment outcomes by credential attainment

```{r outcomes-by-cred}
analysis_flat %>%
  mutate(
    highest_cred = case_when(
      earned_bachelor    == 1 ~ "Bachelor's",
      earned_associate   == 1 ~ "Associate's",
      earned_certificate == 1 ~ "Certificate",
      TRUE                    ~ "No credential"
    ),
    highest_cred = factor(highest_cred,
                          levels = c("No credential", "Certificate",
                                     "Associate's", "Bachelor's"))
  ) %>%
  group_by(highest_cred) %>%
  summarise(
    N                  = n(),
    `Ever employed`    = percent(mean(ever_employed,   na.rm = TRUE), accuracy = 0.1),
    `Stably employed`  = percent(mean(stably_employed, na.rm = TRUE), accuracy = 0.1),
    `Median wages`     = dollar(median(annual_wages_2023[stably_employed == 1],
                                       na.rm = TRUE))
  ) %>%
  kable(caption = "Labor market outcomes by highest credential earned") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 6.4 Employment over time

This chart shows the share of students employed in each quarter over the 
study period. Early quarters are dominated by students still in school; 
employment rates rise as students exit education and enter the workforce.

```{r employment-over-time}
wages_long %>%
  group_by(calendar_quarter, quarter_number) %>%
  summarise(pct_employed = mean(employed, na.rm = TRUE), .groups = "drop") %>%
  arrange(quarter_number) %>%
  mutate(calendar_quarter = factor(calendar_quarter,
                                   levels = unique(calendar_quarter))) %>%
  ggplot(aes(x = calendar_quarter, y = pct_employed, group = 1)) +
  geom_line(color = "#2E86AB", linewidth = 1.1) +
  geom_point(color = "#2E86AB", size = 2) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(
    title = "Share of Cohort Employed by Quarter",
    x     = "Quarter",
    y     = "Share employed"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## 6.5 Wages by sex

Policy analyses routinely examine outcome differences by sex. 
This plot shows wage distributions for stably employed students by sex.

```{r wages-by-sex}
analysis_flat %>%
  filter(stably_employed == 1) %>%
  mutate(sex = fct_reorder(sex,
                                      annual_wages_2023, median, na.rm = TRUE)) %>%
  ggplot(aes(x = sex, y = annual_wages_2023, fill = sex)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.8) +
  scale_y_continuous(labels = dollar_format(scale = 1/1000, suffix = "K")) +
  scale_fill_brewer(palette = "Set2") +
  coord_flip() +
  labs(
    title = "Annual Wage Distribution by Sex",
    subtitle = "Among stably employed students only",
    x     = NULL,
    y     = "Annual wages"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

---

# 7. Next steps

This module covered the structure of the synthetic longitudinal data and 
produced descriptive summaries of demographics, enrollment, credentials, and 
labor market outcomes.

**Module 2 — Sequence Analysis** builds on this by:

- Defining sequences using TraMineR
- Computing pairwise distances between sequences using Optimal Matching
- Visualizing sequence patterns with index plots and state distribution plots

**Module 3 — Cluster Analysis** builds on Module 2 by:

- Applying hierarchical clustering to the distance matrix
- Selecting the number of clusters using statistical fit criteria
- Characterizing and labeling each cluster
- Examining how cluster membership varies by demographics and outcomes


**You have now completed the data preparation and visualization module.** 
No new files were saved in this module. When you are ready to move on, open 
02_sequence_analysis.Rmd to begin the sequence analysis, where you will define 
student sequences, compute pairwise distances, and visualize pathway patterns 
across the cohort.

---

# Session information

The code in this module was developed and tested using the package versions 
listed below. If you encounter unexpected errors, checking your package 
versions against these is a good first troubleshooting step. You can update 
packages by running `update.packages()` in your RStudio console.

```{r session-info}
sessionInfo()
```
**Development Session Info**
R version 4.5.2 (2025-10-31)
Platform: x86_64-pc-linux-gnu
Running under: Red Hat Enterprise Linux 9.8 (Plow)

Matrix products: default
BLAS/LAPACK: FlexiBLAS OPENBLAS-OPENMP;  LAPACK version 3.9.0

locale:
LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8   
LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C           LC_TELEPHONE=C        
LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

time zone: America/New_York
tzcode source: system (glibc)

attached base packages:
stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
kableExtra_1.4.0 knitr_1.50       scales_1.4.0     lubridate_1.9.4  forcats_1.0.1    stringr_1.6.0    dplyr_1.1.4     
purrr_1.2.0      readr_2.1.6      tidyr_1.3.1      tibble_3.3.0     ggplot2_4.0.1    tidyverse_2.0.0 

loaded via a namespace (and not attached):
sass_0.4.10        generics_0.1.4     xml2_1.4.1         stringi_1.8.7      hms_1.1.4          digest_0.6.38     
magrittr_2.0.4     evaluate_1.0.5     grid_4.5.2         timechange_0.3.0   RColorBrewer_1.1-3 fastmap_1.2.0     
jsonlite_2.0.0     viridisLite_0.4.2  jquerylib_0.1.4    textshaping_1.0.4  cli_3.6.5          rlang_1.1.6       
crayon_1.5.3       bit64_4.6.0-1      cachem_1.1.0       withr_3.0.2        yaml_2.3.10        tools_4.5.2       
parallel_4.5.2     tzdb_0.5.0         vctrs_0.6.5        R6_2.6.1           lifecycle_1.0.4    bit_4.6.0         
vroom_1.6.6        pkgconfig_2.0.3    bslib_0.9.0        pillar_1.11.1      gtable_0.3.6       rsconnect_1.6.1   
glue_1.8.0         systemfonts_1.3.1  xfun_0.54          tidyselect_1.2.1   rstudioapi_0.17.1  farver_2.1.2      
htmltools_0.5.8.1  labeling_0.4.3     rmarkdown_2.30     svglite_2.2.2      compiler_4.5.2     S7_0.2.1          

---

*Synthetic data generated for demonstration purposes.
