---
title: "School-to-Work Pathways: Sequence Analysis"
subtitle: "Module 2 of 3 — Using Synthetic Longitudinal Data"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# Overview

This is the second of three modules on studying school-to-work pathways in 
longitudinal administrative data.

This module introduces **sequence analysis**,  which is a method for describing and 
comparing individual trajectories over time. Rather than summarizing outcomes 
at a single point (e.g., "was this person employed at year 6?"), sequence 
analysis treats the entire pathway as the unit of analysis.

## What is a sequence?

A **sequence** is an ordered string of states. In this dataset, each student 
has a sequence of 28 quarters, and each quarter is coded as one of several states:

| State        | Meaning                                    |
|--------------|--------------------------------------------|
| `4YR_Work`   | Enrolled at 4-year college AND employed    |
| `4YR_NoWork` | Enrolled at 4-year college, not employed   |
| `CTC_Work`   | Enrolled at community/technical college AND employed |
| `CTC_NoWork` | Enrolled at community/technical college, not employed |
| `Work_Only`  | Not enrolled, employed                     |
| `Neither`    | Not enrolled, not employed                 |

For example, a student who went directly to a 4-year college and worked part-time 
might have a sequence like:

`4YR_Work → 4YR_Work → 4YR_Work → 4YR_Work → 4YR_Work → 4YR_NoWork → 4YR_NoWork → 4YR_NoWork → Work_Only → Work_Only → ...`

Sequence analysis lets us visualize these patterns, measure how similar two 
sequences are, and eventually cluster students with similar trajectories.

## Key package: TraMineR

We use the **TraMineR** package, developed at the University of Geneva. It 
provides tools for defining, visualizing, and analyzing sequences.

- Documentation: https://traminer.unige.ch
- Key paper: Gabadinho et al. (2011), *Analyzing and Visualizing State 
  Sequences in R with TraMineR*, Journal of Statistical Software
  
  
**Output files**
This module produces the following output files, which are saved to the 
longitudinal_toy_data/ folder and are required for Module 3:

- dist_om.rds — the pairwise Optimal Matching distance matrix used for clustering
- seqdata.rds — the TraMineR sequence object used for visualization


**A note on package versions:** The code in this module was developed and tested 
using specific versions of R and its packages. If you encounter unexpected 
errors or differences in output, it may be due to package version differences. 
See the [Session information] section at the end of this module for the full 
list of package versions used in development. You can check your own package 
versions by running sessionInfo() in your RStudio console, and update packages 
by running update.packages().



---

# 1. Setup

## 1.1 Load packages

```{r load-packages}
library(tidyverse)
library(TraMineR)    # core sequence analysis package
library(scales)
library(knitr)
library(kableExtra)
library(RColorBrewer)

# Install TraMineR if needed:
# install.packages("TraMineR")
```

## 1.2 Load data

```{r load-data}
data_dir <- "/data/share/cdi/Sequences to Success in WA/Quantitative/GitHub Repo/longitudinal_toy_data"

# The combined_seq_wide table has one row per student and one column per quarter.
# Each cell contains the combined enrollment + employment state for that quarter.
combined_wide <- read_csv(file.path(data_dir, "combined_seq_wide.csv"))

# We also load the flat file for demographic comparisons later
analysis_flat <- read_csv(file.path(data_dir, "analysis_flat.csv"))

# And students for demographic variables
students <- read_csv(file.path(data_dir, "students.csv"))
```

## 1.3 Inspect the wide sequence data

```{r inspect-wide}
# The sequence columns are named cstate_1 through cstate_28
# Each row is one student; each column is one quarter
glimpse(combined_wide)

# Preview the first few rows and columns
combined_wide %>%
  select(student_id, cstate_1:cstate_6) %>%
  head(10)
```

---

# 2. Defining sequences in TraMineR

## 2.1 Specify the alphabet and labels

Before creating sequence objects, we define the **alphabet**, or the complete 
set of possible states. TraMineR uses this to validate the data and produce 
consistent visualizations.

```{r define-alphabet}
# The alphabet: all possible states a student can be in at any given quarter
STATES <- c(
  "4YR_Work",   # 4-year + employed
  "4YR_NoWork", # 4-year, not employed
  "CTC_Work",   # CTC + employed
  "CTC_NoWork", # CTC, not employed
  "Work_Only",  # workforce only
  "Neither"     # neither enrolled nor employed
)

# Human-readable labels for plots (same order as STATES)
LABELS <- c(
  "4-Year + Work",
  "4-Year Only",
  "CTC + Work",
  "CTC Only",
  "Work Only",
  "Neither"
)

# Color palette — one color per state
# Using a manually chosen palette for visual clarity
STATE_COLORS <- c(
  "4YR_Work"   = "#1B6CA8",  # deep blue
  "4YR_NoWork" = "#6EB5E0",  # light blue
  "CTC_Work"   = "#7B2D8B",  # deep purple
  "CTC_NoWork" = "#C490D1",  # light purple
  "Work_Only"  = "#2CA25F",  # green
  "Neither"    = "#CCCCCC"   # gray
)
```

## 2.2 Create the sequence object

The `seqdef()` function converts the wide data frame into a TraMineR 
**sequence object** which is a special data structure that stores sequences alongside 
metadata like the alphabet, labels, and colors.

```{r create-seqobj}
# Extract just the sequence columns (cstate_1 through cstate_28)
seq_cols <- grep("^cstate_", names(combined_wide), value = TRUE)

# Create the TraMineR sequence object
# - The first argument is the data frame containing sequence columns
# - states = the alphabet (all possible values)
# - labels = human-readable names for each state
# - cnames = names for the time points (here: quarter 1 through quarter 28)
# - xtstep = how often to label the x-axis in plots
seqdata <- seqdef(
  data     = combined_wide[, seq_cols],
  alphabet = STATES,
  states   = STATES,
  labels   = LABELS,
  cnames   = paste0("Quarter ", 1:28),
  cpal     = unname(STATE_COLORS[STATES]),
  xtstep   = 4,
  missing.color = "#F5C842"
)

# Inspect the sequence object
options(max.print = 20) # limit rows that will be printed
print(seqdata,  format = "SPS")  # SPS = Successive Periods of States format
```

```{r seqobj-summary}
# seqdef produces a sequence object — check its basic properties
cat("Number of sequences:", nrow(seqdata), "\n")
cat("Sequence length:    ", ncol(seqdata), "quarters\n")
cat("Alphabet:           ", paste(alphabet(seqdata), collapse = ", "), "\n")
```

---

# 3. Visualizing sequences

## 3.1 State distribution plot (tempogram)

The **state distribution plot** shows the proportion of students in each state 
at each time point. It is the most common way to get an overall sense of how 
the cohort moves through states over time. Each vertical slice represents one 
quarter, and the colored bands show what proportion of the cohort was in each 
state at that point in time. The height of each band corresponds to the share 
of students in that state, and all bands together add up to 100% in every 
quarter.

**Example Interpretation (Remember Data Are Synthetic)**
In the earliest quarters, a large share of 
students are enrolled in either a 4-year college (light blue) or a CTC (pink and
purple), with enrollment peaking around Quarter 5. Notably, a meaningful portion
of enrolled students are also working at the same time (the darker blue and 
purple bands), reflecting the reality that many students balance school and 
employment simultaneously.

As the cohort moves through time, enrollment tapers off as students complete 
their programs or stop out. By the later quarters, the chart is dominated by two
states: Work Only (green), representing students who have finished school and 
entered the workforce, and Neither (gray), representing students who are neither
enrolled nor employed at that point.

The growth of the Work Only band over time is an encouraging pattern as it 
suggests that as students complete their pathways, a growing share transition 
into stable employment. The persistence of the Neither band in later quarters, 
however, points to a group of students who may need additional support 
connecting to the labor market.

```{r state-dist-plot, fig.height=5}
# seqdplot() = state distribution plot
# withlegend = "right" places the legend to the right of the plot
# border = NA removes cell borders for a cleaner look
seqdplot(
  seqdata,
  title      = "State Distribution by Quarter (Full Cohort)",
  withlegend = "right",
  border     = NA,
  ylab       = "Proportion of students",
  xlab       = "Quarter"
)


```



## 3.2 Sequence index plot

The **sequence index plot** (also called an index plot or "carpet plot") shows 
one row per student, with each row colored by state at each time point. Each 
horizontal line represents one student, and the colors show which state that 
student was in at each quarter across the full 28-quarter observation window. 
Students are sorted by their starting state, so similar pathways tend to appear 
grouped together. With 5,000 students this is dense, but sorting by sequence 
pattern makes structure visible.

**Example Interpretation (Remember Data Are Synthetic)**
Several distinct pathway patterns are visible. At the bottom of the chart, a 
group of students (shown in dark and light blue) start in 4-year college 
enrollment, many of whom transition into Work Only (green) after completing 
their degrees. In the middle, a band of students follow CTC pathways (purple 
shades) before also moving into the workforce. Toward the top, a large group of
students show gray (Neither) or green (Work Only) from early on, reflecting 
those who never enrolled in postsecondary education or who moved directly into
the labor market.

The increasing dominance of green (Work Only) in the later quarters is visible 
across most pathway groups, reinforcing the pattern seen in the state 
distribution plot. Over time, more and more students transition into employment 
regardless of their starting pathway.

The variation and complexity of the individual lines illustrates exactly why 
sequence analysis is useful. Rather than summarizing outcomes at a single point 
in time, it captures the full trajectory of each student, making it possible to 
identify groups of students who followed meaningfully similar paths. That 
grouping is what the cluster analysis in the next module will accomplish.


```{r index-plot-full, fig.height=7}
# seqiplot() = sequence index plot
# sortv = "from.start" sorts sequences by their starting state,
#         which groups similar trajectories together visually
# idxs = 0 means "plot all sequences" (use a number like 500 for a sample)
seqiplot(
  seqdata,
  title      = "Sequence Index Plot — All Students (sorted by start state)",
  sortv      = "from.start",
  withlegend = "right",
  border     = NA,
  ylab       = "Students",
  xlab       = "Quarter",
  idxs       = 0
)
```



```{r index-plot-sample, fig.height=6}
# With 5,000 students the full plot is dense.
# A random sample of 100 is easier to read and still shows the main patterns.
set.seed(25)
sample_idx <- sample(nrow(seqdata), 100)

seqiplot(
  seqdata[sample_idx, ],
  title      = "Sequence Index Plot — Random Sample of 100 Students",
  sortv      = "from.start",
  withlegend = "right",
  border     = NA,
  ylab       = "Students",
  xlab       = "Quarter",
  idxs       = 1:100
)
```

## 3.3 Sequence frequency plot

The **sequence frequency plot** shows only the most common distinct sequences, 
ranked by frequency. This is useful for identifying dominant pathway patterns.

```{r freq-plot, fig.height=6}
# seqfplot() = sequence frequency plot
# idxs = 1:20 shows the 20 most frequent sequences
seqfplot(
  seqdata,
  title      = "20 Most Frequent Sequences",
  idxs       = 1:20,
  withlegend = "right",
  border     = NA,
  ylab       = "Cumulative frequency (%)",
  xlab       = "Quarter"
)
```

## 3.4 Mean time spent in each state

The `seqmtplot()` function shows the average number of quarters each student 
spends in each state. This is a simple but useful summary statistic.

**Example Interpretation (Remember Data Are Synthetic)**
This bar chart shows the average number of quarters students spent in each state 
across the full 28-quarter observation window. The two most time-consuming 
states are Work Only (about 9 quarters on average) and Neither (about 8 
quarters), reflecting the fact that most students spend the majority of the 
observation period outside of enrollment. Enrollment states, both with and 
without concurrent employment, account for a relatively smaller share of the 
average student's trajectory, typically 3 to 4 quarters each for 4-year and CTC 
pathways. This reinforces that while postsecondary enrollment is an important 
part of the pathway, much of a student's time in the observation window is spent
in the labor market or disconnected from both school and work.

```{r mean-time-plot, fig.height=4.5}
seqmtplot(
  seqdata,
  title      = "Mean Time Spent in Each State",
  withlegend = "right",
  ylab       = "Mean number of quarters"
)
```

---

# 4. Sequence statistics

## 4.1 Transition rates

**Transition rates** describe how likely students are to move from one state 
to another between consecutive quarters. `seqtrate()` returns a matrix of 
state-to-state transition probabilities. Each row represents a "from" state and 
each column represents a "to" state. The values show the probability that a 
student in a given state in one quarter will be in each possible state the 
following quarter. High values on the diagonal indicate that students tend to 
stay in the same state from one quarter to the next Values in each row sum to 1.

**Example Interpretation (Remember Data Are Synthetic)**
Enrollment states are moderately persistent but show realistic movement. 
Students enrolled at a 4-year college and working (4YR_Work) have a 58% chance 
of staying in that same state the next quarter, but also have meaningful 
probabilities of dropping to 4YR_NoWork (22%) or transitioning out of enrollment 
entirely into Work Only (11%). Similarly, CTC_Work students have a 54% chance of 
staying, with notable transitions to CTC_NoWork (21%) and Work Only (14%). This 
suggests that students frequently shift between working and not working while 
enrolled.

Students not working while enrolled often stop out. 4YR_NoWork students have a 
68% chance of staying in that state, but an 11% chance of stopping out to 
Neither. This is higher than the stop-out rate for working students. A similar 
pattern holds for CTC_NoWork students, who have a 15% chance of transitioning to 
Neither. This suggests that working while enrolled may be associated with 
greater persistence, though this is a pattern in the synthetic data and should 
be interpreted cautiously.

Work Only and Neither are fairly sticky. Students in Work Only stay there 75% of 
the time, while students in Neither stay there 71% of the time. However, Neither 
students do show a 24% transition into Work Only, reflecting students gradually 
connecting to the labor market over time.

```{r transition-rates}
# seqtrate() computes the empirical transition rate matrix
trate <- seqtrate(seqdata)

# Round for readability and display as a table
round(trate, 3) %>%
  as.data.frame() %>%
  rownames_to_column("From \\ To") %>%
  kable(caption = "Transition rates between states (row = from, column = to)") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                font_size = 11, full_width = TRUE) %>%
  column_spec(1, bold = TRUE)
```



## 4.2 Sequence entropy

**Entropy** measures how unpredictable a sequence is. A sequence that stays in 
one state has entropy = 0; a sequence that jumps randomly between all states 
has high entropy. In pathway research, high entropy may indicate students with 
complex or non-linear trajectories.

**Example Interpretation (Remember Data Are Synthetic)**
The distribution has two notable features. First, there is a small cluster of 
students at or near 0, representing students with very simple, uniform pathways
who most likely spent nearly all 28 quarters in Neither or Work Only without 
transitioning. Second, the bulk of the distribution is spread across entropy 
values of roughly 0.35 to 0.90, with a peak around 0.40 and a second 
concentration around 0.65 to 0.80. This spread reflects the genuine diversity of 
pathways in the cohort with some students following relatively straightforward 
trajectories while others move through multiple states over time.

The bimodal shape, with one concentration around 0.40 and another around 0.65 to
0.80, hints that there may be meaningful subgroups in the data with 
systematically different levels of pathway complexity. This is exactly the kind 
of variation that sequence and cluster analysis is designed to uncover. This
pattern also suggests that the cluster analysis in the next module is likely to 
identify distinct groups of students with meaningfully different trajectories.

```{r entropy}
# seqient() computes the entropy of each individual sequence
entropy <- seqient(seqdata)

# Plot distribution of entropy across students
data.frame(entropy = as.numeric(entropy)) %>%
  ggplot(aes(x = entropy)) +
  geom_histogram(binwidth = 0.02, fill = "#2E86AB", color = "white", alpha = 0.85) +
  labs(
    title    = "Distribution of Sequence Entropy",
    subtitle = "Higher values indicate more complex, variable pathways",
    x        = "Sequence entropy",
    y        = "Number of students"
  ) +
  theme_minimal(base_size = 13)
```

## 4.3 Number of transitions

`seqtransn()` counts the number of state changes in each sequence. A student 
who goes CTC → Work has 1 transition; a student who enrolls, leaves, and 
re-enrolls has more.

**Example Interpretation (Remember Data Are Synthetic)**
The distribution is roughly bell-shaped and centered around 7 to 9 transitions, 
meaning the typical student changed states multiple times over the observation 
period. This is consistent with the transition rate table, which showed that 
most states have meaningful probabilities of moving to other states from quarter 
to quarter.

The small group of students with 0 or 1 transitions represents those with very 
stable, uniform pathways. On the other end, a small number of students made 13 
or more transitions, reflecting highly variable trajectories with frequent 
movement between states.

Overall, the relatively high number of transitions per student reinforces that 
these pathways are genuinely complex and dynamic.

```{r transitions}
n_transitions <- seqtransn(seqdata)

data.frame(n_trans = as.numeric(n_transitions)) %>%
  ggplot(aes(x = n_trans)) +
  geom_bar(fill = "#A23B72", alpha = 0.85) +
  scale_x_continuous(breaks = 0:10) +
  labs(
    title = "Number of State Transitions per Student",
    x     = "Number of transitions",
    y     = "Number of students"
  ) +
  theme_minimal(base_size = 13)
```

---

# 5. Subgroup comparisons

One of the most useful things you can do with sequence analysis is compare 
trajectory patterns across demographic subgroups. Here we compare state 
distributions by FRPL status and by rural/urban.

## 5.1 By FRPL status
These two charts show state distribution plots side by side, comparing students 
who received free or reduced price lunch (FRPL) in high school (a common 
indicator of economic disadvantage) with those who did not. The charts follow 
the same format as the full cohort state distribution plot above, with each 
vertical slice showing the proportion of students in each state at a given 
quarter.

**Example Interpretation (Remember Data Are Synthetic)**
Several meaningful differences stand out between the two groups. Non-FRPL 
students show substantially higher rates of 4-year college enrollment (light 
blue and dark blue) in the early quarters, peaking around Quarter 5 before 
tapering off as students complete their degrees. FRPL students, by contrast, 
show lower 4-year enrollment throughout and a relatively higher share in CTC 
pathways (purple shades), which may reflect differences in postsecondary access 
and choice by income level.

Both groups show a similar overall arc, with enrollment concentrated in the 
early quarters and Work Only growing over time, but the transition happens 
somewhat differently. Non-FRPL students show a more pronounced shift into Work 
Only in the middle quarters, consistent with more students completing bachelor's 
degrees on a "normative" timeline. FRPL students show a higher proportion in 
Neither across most of the observation window, which may suggest greater 
challenges connecting to both postsecondary education and the labor market.

```{r subgroup-frpl, fig.height=5}
# Merge FRPL indicator onto the sequence data index
frpl_vec <- students$frpl[match(combined_wide$student_id, students$student_id)]
frpl_label <- if_else(frpl_vec == 1, "FRPL", "Non-FRPL")

# seqdplot with group argument produces side-by-side distribution plots
seqdplot(
  seqdata,
  group      = frpl_label,
  title      = "State Distribution by FRPL Status",
  withlegend = "right",
  border     = NA
)
```

## 5.2 By rural/urban
These two charts compare state distributions over time for students from 
non-rural and rural areas. The format is the same as the previous subgroup 
comparisons, with each vertical slice showing the proportion of students in each 
state at a given quarter.

**Example Interpretation (Remember Data Are Synthetic)**
The differences between rural and non-rural students are notable, though 
somewhat less pronounced than the FRPL differences above. Non-rural students 
show higher rates of 4-year college enrollment (blue shades) in the early 
quarters, with a clear peak around Quarter 5 before declining as students 
complete their programs. Rural students show lower 4-year enrollment throughout 
the observation window and a somewhat higher share in CTC pathways (purple 
shades), consistent with patterns seen in research on rural postsecondary access 
where geographic distance and economic factors often make community and 
technical colleges a more accessible option.

Both groups transition toward Work Only (green) over time, but rural students 
show a persistently higher share in Neither across the observation window, 
suggesting that connecting to the labor market may be more challenging for 
students from rural areas. The lower overall enrollment rates for rural students 
in the early quarters also point to a larger share going directly into the 
workforce or neither working nor enrolling after high school graduation.

As with the FRPL comparison, these differences illustrate how pathway patterns 
can vary meaningfully across student subgroups, and how sequence analysis makes 
those differences visible in a way that single-point outcome measures cannot.

```{r subgroup-rural, fig.height=5}
rural_vec   <- students$rural_flag[match(combined_wide$student_id, students$student_id)]
rural_label <- if_else(rural_vec == 1, "Rural", "Non-Rural")

seqdplot(
  seqdata,
  group      = rural_label,
  title      = "State Distribution by Rural/Urban",
  withlegend = "right",
  border     = NA
)
```


# 6. Computing pairwise distances (Optimal Matching)

The core computational step in sequence analysis is calculating a **distance 
matrix** — a measure of how different each pair of sequences is from each other.

## 6.1 What is Optimal Matching?

**Optimal Matching (OM)** is borrowed from bioinformatics (DNA sequence 
alignment). It finds the minimum "cost" to transform one sequence into another 
using insertions, deletions, and substitutions. Two sequences are "close" if 
they require few operations to make them identical.

Key parameters:

- **Indel cost:** The cost of inserting or deleting a state. Higher indel 
  costs penalize timing differences more.
- **Substitution cost matrix:** The cost of swapping one state for another. 
  We use `sm = "TRATE"` to derive substitution costs from empirical transition 
  rates — states that rarely transition to each other get a high substitution 
  cost.

## 6.2 Compute the distance matrix

```{r compute-distances}
# This step can take a minute or two with 5,000 sequences.
# The result is an N x N distance matrix (stored as a "dist" object).

# sm = "TRATE" derives substitution costs from observed transition rates.
#   This is the most common approach for social science sequence data.
# indel = 1.5 is a standard starting value; you can experiment with this.
# norm = "auto" normalizes distances by sequence length (important when
#   sequences have different lengths due to missing data)

cat("Computing pairwise distances using Optimal Matching...\n")

dist_om <- seqdist(
  seqdata,
  method = "OM",
  sm     = "TRATE",
  indel  = 1.5,
  norm   = "auto",
  with.missing = TRUE
)

cat("Distance matrix dimensions:", dim(dist_om), "\n")
cat("Range of distances:", round(range(dist_om), 3), "\n")
```

**Example Interpretation (Remember Data Are Synthetic)**
This graph below shows the distribution of pairwise Optimal Matching (OM) 
distances across all possible pairs of students in the cohort. Each value 
represents how dissimilar two students' sequences are, with 0 meaning identical 
pathways and higher values indicating greater dissimilarity. The distribution is 
roughly bell-shaped and centered around 0.7 to 0.8, meaning most pairs of 
students have moderately to highly dissimilar trajectories. The small cluster of 
values near 0 represents pairs of students whose pathways are nearly identical, 
likely students who both spent the full observation window in Neither or Work 
Only. The overall spread of the distribution confirms that there is substantial 
variation in pathways across the cohort, which is a good sign that the cluster 
analysis will be able to identify meaningfully distinct groups.

```{r distance-dist}
# The distribution of pairwise distances tells you how spread out sequences are.
# A bimodal distribution often suggests distinct pathway clusters.
data.frame(dist = as.vector(dist_om[lower.tri(dist_om)])) %>%
  ggplot(aes(x = dist)) +
  geom_histogram(bins = 60, fill = "#2E86AB", color = "white", alpha = 0.85) +
  labs(
    title    = "Distribution of Pairwise Optimal Matching Distances",
    subtitle = "Each value represents the dissimilarity between two student sequences",
    x        = "OM distance",
    y        = "Frequency"
  ) +
  theme_minimal(base_size = 13)
```

## 6.3 A note on alternative distance measures
OM is the most widely used method, but TraMineR supports others. The right 
choice depends on what aspect of the pathway you care most about. The table 
below describes each method and provides examples of research questions where it 
would be a good fit.

* OM (Optimal Matching): Counts the minimum number of insertions, deletions, and
substitutions needed to turn one sequence into another. This is the general 
purpose method that works well when both the states visited and their timing 
matter. A good fit for questions like: How do school-to-work pathways differ 
between students who received FRPL and those who did not?

*HAM (Hamming distance): Compares sequences position by position and counts 
mismatches. Works best when sequences are the same length and timing is 
important. It is also computationally faster than OM. A good fit for questions 
like: Do students from rural areas differ from urban students in their 
enrollment status in each quarter of the first two years after high school?

*LCS (Longest Common Subsequence): Measures similarity based on shared states 
appearing in the same order, regardless of when they occur. A good fit for 
questions like: Do students who eventually transfer from a CTC to a 4-year 
college share a common sequence of early pathway steps, even if the timing 
varies?

*DHD (Dynamic Hamming Distance): A time-sensitive version of Hamming that allows 
for some flexibility in timing. A good fit for questions like: Are there groups 
of students who follow similar enrollment patterns in the first year after high 
school, even if some start a quarter earlier or later than others?

*OMslen (OM with spell length weighting): Modifies OM to give more weight to 
time spent in a state rather than just the transitions between states. A good 
fit for questions like: Does the length of time a student spends enrolled at a 
CTC before transferring predict their likelihood of completing a bachelor's 
degree?

For policy research on education and workforce pathways, OM with TRATE 
substitution costs is a reasonable default. It is flexible, well-understood, and
produces results that are interpretable for a broad audience. The other methods 
are worth exploring if you have a specific hypothesis about whether timing, 
duration, or sequence order is the most important dimension of the pathway you 
are studying.

---

# 7. Saving the distance matrix

The distance matrix is the main output of sequence analysis and the main input 
to cluster analysis. We save it here so Module 3 can load it directly without 
recomputing.

```{r save-distances}
# Save as an R data file (.rds) — this preserves the matrix class and attributes
saveRDS(dist_om,  file = file.path(data_dir, "dist_om.rds"))
saveRDS(seqdata,  file = file.path(data_dir, "seqdata.rds"))

cat("Saved dist_om.rds and seqdata.rds to", data_dir, "\n")
```

---

# 8. Summary

In this module we:

1. Defined sequences using TraMineR's `seqdef()` function
2. Visualized the full cohort using state distribution plots, index plots, 
   and frequency plots
3. Computed sequence statistics including transition rates, entropy, and 
   number of transitions
4. Compared trajectory patterns across demographic subgroups
5. Computed a pairwise Optimal Matching distance matrix

**Module 3 — Cluster Analysis** will take the distance matrix computed here 
and use hierarchical clustering to group students with similar trajectories 
into pathway clusters. We will then label those clusters, validate them, and 
examine how they vary by demographics and outcomes.

**You have now completed the sequence analysis module.** 
The following files have been saved to your longitudinal_toy_data/ folder and 
will be loaded automatically at the start of Module 3:
- dist_om.rds — the pairwise distance matrix computed in Section 6
- seqdata.rds — the TraMineR sequence object used for visualization

Open 03_cluster_analysis.Rmd to continue to the next step, where you will use 
these files to group students with similar trajectories into pathway clusters.

---

# Session information

The code in this module was developed and tested using the package versions 
listed below. If you encounter unexpected errors, checking your package 
versions against these is a good first troubleshooting step. You can update 
packages by running `update.packages()` in your RStudio console.

```{r session-info}
sessionInfo()
```


**Development Session Info**
R version 4.5.2 (2025-10-31)
Platform: x86_64-pc-linux-gnu
Running under: Red Hat Enterprise Linux 9.8 (Plow)

Matrix products: default
BLAS/LAPACK: FlexiBLAS OPENBLAS-OPENMP;  LAPACK version 3.9.0

locale:
LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8   
LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C           LC_TELEPHONE=C        
LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

time zone: America/New_York
tzcode source: system (glibc)

attached base packages:
stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
RColorBrewer_1.1-3 TraMineR_2.2-12    kableExtra_1.4.0   knitr_1.50         scales_1.4.0       lubridate_1.9.4   
forcats_1.0.1      stringr_1.6.0      dplyr_1.1.4        purrr_1.2.0        readr_2.1.6        tidyr_1.3.1       
tibble_3.3.0       ggplot2_4.0.1      tidyverse_2.0.0   

loaded via a namespace (and not attached):
gtable_0.3.6      xfun_0.54         bslib_0.9.0       lattice_0.22-7    tzdb_0.5.0        vctrs_0.6.5      
tools_4.5.2       generics_0.1.4    parallel_4.5.2    cluster_2.1.8.1   pkgconfig_2.0.3   Matrix_1.7-4     
S7_0.2.1          lifecycle_1.0.4   compiler_4.5.2    farver_2.1.2      textshaping_1.0.4 permute_0.9-8    
htmltools_0.5.8.1 sass_0.4.10       yaml_2.3.10       pillar_1.11.1     crayon_1.5.3      jquerylib_0.1.4  
MASS_7.3-65       rsconnect_1.6.1   cachem_1.1.0      vegan_2.7-2       boot_1.3-32       nlme_3.1-168     
tidyselect_1.2.1  digest_0.6.38     stringi_1.8.7     splines_4.5.2     labeling_0.4.3    fastmap_1.2.0    
grid_4.5.2        colorspace_2.1-2  cli_3.6.5         magrittr_2.0.4    withr_3.0.2       bit64_4.6.0-1    
timechange_0.3.0  rmarkdown_2.30    bit_4.6.0         hms_1.1.4         evaluate_1.0.5    viridisLite_0.4.2
mgcv_1.9-3        rlang_1.1.6       glue_1.8.0        xml2_1.4.1        svglite_2.2.2     rstudioapi_0.17.1
vroom_1.6.6       jsonlite_2.0.0    R6_2.6.1          systemfonts_1.3.1

---

*Synthetic data generated for teaching purposes.*
