---
title: "School-to-Work Pathways: Cluster Analysis"
subtitle: "Module 3 of 3 — Using Synthetic Longitudinal Data"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
```

---

# Overview

This is the third and final module on studying school-to-work pathways in 
longitudinal administrative data.

In Module 2, we computed a pairwise **Optimal Matching distance matrix** that 
captures how similar or different each pair of student sequences is. In this 
module we use that distance matrix as the input to **cluster analysis** which is 
a method for grouping students with similar trajectories into meaningful pathway 
clusters.

This module covers:

1. Choosing a clustering method
2. Determining the number of clusters
3. Running the final clustering
4. Visualizing and labeling clusters
5. Validating clusters using demographics and outcomes
6. Examining disparities in pathway participation


**Output files**
This module produces the following output file, which is saved to the 
longitudinal_toy_data/ folder:

-final_results_with_clusters.csv — one row per student with demographic 
characteristics, high school records, credential attainment, labor market 
outcomes, and pathway cluster membership merged into a single analysis-ready file.


**A note on package versions:** The code in this module was developed and tested 
using specific versions of R and its packages. If you encounter unexpected 
errors or differences in output, it may be due to package version differences. 
See the [Session information] section at the end of this module for the full 
list of package versions used in development. You can check your own package 
versions by running sessionInfo() in your RStudio console, and update packages 
by running update.packages().

---

# 1. Setup

## 1.1 Load packages

```{r load-packages}
library(tidyverse)
library(TraMineR)        # sequence visualization
library(cluster)         # agnes() for hierarchical clustering
library(WeightedCluster) # cluster quality statistics and tools
library(scales)
library(knitr)
library(kableExtra)
library(ggridges)        # for ridge/density plots by group

# Install any missing packages:
#install.packages(c("cluster", "WeightedCluster", "ggridges"))

```

## 1.2 Load data

```{r load-data}
data_dir <- "synthetic_data"

# Load the distance matrix and sequence object saved in Module 2
dist_om  <- readRDS(file.path(data_dir, "dist_om.rds"))
seqdata  <- readRDS(file.path(data_dir, "seqdata.rds"))

# Load the flat analysis file and student demographics
analysis_flat  <- read_csv(file.path(data_dir, "analysis_flat.csv"))
students       <- read_csv(file.path(data_dir, "students.csv"))
combined_wide  <- read_csv(file.path(data_dir, "combined_seq_wide.csv"))
```

---

# 2. Choosing a clustering method

## 2.1 Why hierarchical clustering?

There are many clustering methods available. For sequence analysis in social 
science, **agglomerative hierarchical clustering** with Ward's linkage criterion 
is by far the most common choice. Here is why:

- **No need to prespecify k:** Hierarchical clustering builds a full tree 
  (dendrogram) from which you can cut at any level. This lets you explore 
  different numbers of clusters before committing.
- **Ward's linkage:** At each step, it merges the two clusters whose merger 
  minimizes the increase in total within-cluster variance. This tends to produce 
  compact, balanced clusters that are desirable for pathway research.
- **Works with a distance matrix:** Because we already have a precomputed 
  distance matrix from Optimal Matching, we can feed it directly into the 
  clustering algorithm.

Alternative methods you might encounter:

* agnes(ward): Uses Ward linkage, the most common approach for OM distances. 
Tends to produce compact, evenly sized clusters.

*agnes(complete): Uses complete linkage. Tends to produce unequal cluster sizes, 
with some clusters much larger than others.

*agnes(average): Uses average linkage. More sensitive to outliers than Ward or 
complete linkage.

*k-medoids (pam): A good alternative to hierarchical clustering that can produce 
more robust results. Worth considering if your clusters are not well-separated, 
though it runs more slowly on large datasets.



## 2.2 Run the hierarchical clustering

```{r run-clustering}
# agnes() from the cluster package fits agglomerative hierarchical clustering.
# method = "ward" uses Ward's minimum variance linkage.
# diss = TRUE tells it that the input is a dissimilarity/distance matrix,
#   not raw data.

cat("Running hierarchical clustering (Ward linkage)...\n")

hc <- agnes(
  x      = dist_om,
  diss   = TRUE,
  method = "ward"
)


cat("Clustering complete.\n")
cat("Agglomerative coefficient:", round(hc$ac, 3),
    "  (closer to 1 = stronger cluster structure)\n")
```

## 2.3 Visualize the dendrogram

The **dendrogram** shows the full hierarchical tree.  Each leaf at the bottom 
represents a student, and the branches show how students are progressively 
grouped together into clusters based on the similarity of their sequences. The
height of each branch (shown on the y-axis as OM distance) indicates how 
dissimilar the groups being merged are at that point. Branches that merge low on 
the chart indicate students with very similar pathways, while branches that 
merge high indicate groups that are quite different from each other.
The three dashed horizontal lines show where the dendrogram would be "cut" to 
produce different numbers of clusters. Cutting at the teal line (k = 5) produces 
5 clusters, cutting at the orange line (k = 8) produces 8 clusters, and cutting 
at the red line (k = 12) produces 12 clusters. The higher the cut line, the
fewer and broader the clusters; the lower the cut line, the more and finer the 
clusters.

**Example Interpretation (Remember Data Are Synthetic)**
The dendrogram shows two very broad groupings at the highest level, reflecting a 
fundamental divide in the cohort that likely separates students who enrolled in 
postsecondary education from those who did not. Within each of those broad 
groups, finer subdivisions emerge as the cut line moves lower. The relatively 
large jump in height before the final merge at the top of the chart suggests 
that a solution with fewer than 12 clusters may be sufficient to capture the 
main pathway patterns without overfitting. The cluster fit statistics in the 
next section will help determine the most appropriate number of clusters for 
this data.

```{r dendrogram, fig.height=5}
# pltree() plots the hierarchical clustering tree
# We only show the top portion (hang = -1, labels = FALSE) since 5,000 
# individual labels would be unreadable
pltree(
  hc,
  main   = "Hierarchical Clustering Dendrogram (Ward Linkage)",
  labels = FALSE,
  hang   = -1,
  ylab   = "Height (OM distance)"
)

# Add reference lines at several potential cut points
abline(h = c(3, 4, 5), col = c("#E84855", "#F4A261", "#2CA25F"),
       lty = 2, lwd = 1.5)
legend("topright",
       legend = c("k ≈ 12 cut", "k ≈ 8 cut", "k ≈ 5 cut"),
       col    = c("#E84855", "#F4A261", "#2CA25F"),
       lty    = 2, lwd = 1.5, cex = 0.9)
```

---

# 3. Choosing the number of clusters

## 3.1 Statistical quality criteria

The `WeightedCluster` package provides a suite of cluster quality statistics 
that help evaluate different values of k. We compute these across a range of k 
values and look for the one that optimizes the criteria.

Key statistics:

- **ASW (Average Silhouette Width):** How well each observation fits its cluster 
versus the next-best alternative. Higher is better, with a maximum of 1.

- **CH (Calinski-Harabasz index):** Ratio of between-cluster to within-cluster 
variance. Higher is better.

- **R2 (R-squared):** Proportion of total variance explained by the cluster 
solution. Higher is better.

- **HG (Hubert's Gamma):** Rank correlation between distances and cluster 
membership. Higher is better.


None of these is definitive, so use them alongside domain knowledge and the 
interpretability of resulting clusters.



The chart below shows four cluster quality statistics (ASW, CH, R2, and HG) 
plotted across a range of possible cluster solutions from 2 to 15 clusters. The 
statistics have been standardized so they can be displayed on the same scale, 
with higher values indicating a better-fitting solution for all four measures. 
The goal is to identify a number of clusters where the statistics stabilize or 
show a notable peak, similar to looking for an "elbow" in the line.


**Example Interpretation (Remember Data Are Synthetic)**
All four statistics show their highest values at k = 2, meaning a two-cluster 
solution fits the data best in a purely statistical sense. However, a 
two-cluster solution is rarely useful for policy research because it likely just 
separates students who enrolled in postsecondary education from those who did 
not, without capturing the meaningful variation within those broad groups.
Looking further along the x-axis, the statistics generally decline and begin to 
flatten out around k = 8 to 10, with no strong peaks emerging at higher values. 
R2 and HG show a slight upward trend continuing beyond k = 10, suggesting that 
adding more clusters continues to explain additional variation, though with 
diminishing returns.

In practice, choosing a cluster solution requires balancing statistical fit with 
interpretability. A solution in the range of k = 8 to 12 is worth exploring, as 
it is large enough to capture meaningful pathway distinctions while remaining 
manageable to interpret and communicate to a policy audience. The sequence index 
plots by cluster (see below) and your substantive knowledge should guide the 
final decision.
```{r cluster-quality, fig.height=7}
# as.clustrange() computes quality statistics for k = 2 through ncluster
# This can take a moment on large datasets
wc_stats <- as.clustrange(
  object   = hc,
  diss     = dist_om,
  ncluster = 15    # evaluate k = 2 through 15
)

# Plot all quality statistics across k values
# Each panel shows a different criterion; look for peaks or elbows
plot(
  wc_stats,
  stat     = c("ASW", "CH", "R2", "HG"),
  norm     = "zscore",   # z-score normalizes criteria to the same scale
  main     = "Cluster Quality Statistics by Number of Clusters"
)
```

This table shows the top 3 cluster solutions for each of 10 quality criteria, 
ranked from best to worst within each criterion. For each criterion, the table 
shows the number of clusters (N groups) and the corresponding statistic value. 
Higher values indicate better fit for all criteria shown.

**Example Interpretation (Remember Data Are Synthetic)**
The results are mixed, which is typical for cluster analysis. A few patterns 
stand out:
-Several criteria (ASW, ASWw, CH, CHsq, and PBC) favor smaller solutions in the 
range of k = 2 to 5, reflecting the fact that a small number of broad groups 
explains a large share of the overall variation. This is statistically sensible 
but may not be substantively useful for policy research.

-On the other hand, HG, HGSD, R2, R2sq, and HC consistently favor larger 
solutions in the range of k = 13 to 15, suggesting that finer distinctions 
between pathways continue to improve fit even at higher values of k.

The lack of consensus across criteria is actually useful information. 
It suggests that there is no single objectively correct number of clusters and 
that the final decision should be guided by what is most meaningful and 
interpretable for your research question. For demonstration purposes, we have 
selected a 12-cluster solution to align with the findings from the Sequences to
Success Project. 
```{r quality-table}
# Print the quality statistics as a table for closer inspection
summary(wc_stats, max.rank = 3) %>%
  kable(caption = "Top 3 cluster solutions by each quality criterion") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## 3.2 Practical considerations

Statistical criteria help narrow the choice, but the final decision should 
also consider:

- **Interpretability:** Do the clusters correspond to meaningfully different 
  pathways that can be labeled and communicated to a policy audience?
  
- **Cluster size:** Very small clusters (e.g., < 1% of the sample) may not 
  be substantively meaningful or statistically stable.
  
- **Research question:** If you are trying to recover a known typology 
  (as we are here), prior theory should guide k.
  
- **Stability:** Run the clustering on a bootstrap sample and check whether 
  the same clusters emerge.

For this synthetic dataset, **k = 12** is the known number of pathway archetypes. 
We will use that as our target.


**Things to Check for When Comparing Different Values of k:** 
-Which pathways get merged together when k is smaller? 

-Does the reduced granularity change any substantive conclusions about earnings 
differences across pathways?

-Are any clusters very small (e.g., less than 2-3% of the sample)? Very small 
clusters may reflect outlier sequences rather than a meaningful pathway group 
and could be a sign that k is too large.

-Are any clusters very large (e.g., more than 30-40% of the sample)? Dominant 
clusters may be masking important variation and could be a sign that k 
is too small.

-Do the pathway labels still make substantive sense? A solution that is 
statistically optimal but produces clusters that are difficult to name or 
explain to a policy audience may not be the most useful choice.

-How stable are the clusters? If moving from k to k+1 causes many students to 
shift clusters dramatically, that may indicate the solution is sensitive to the 
choice of k.

-Do subgroup patterns hold across different values of k? For example, if FRPL 
students are overrepresented in a particular pathway at k=12, does that pattern 
persist at k=8 or k=10?

-Does the solution align reasonably with what you know about the population? 

---

# 4. Assigning cluster membership
Once you have selected a cluster solution, the final step is to assign each 
student to a cluster. The cutree() function cuts the dendrogram at your chosen 
value of k and returns a cluster membership vector with one value per student. 
This vector is then added to the data as a new variable so that cluster 
membership can be used in subsequent analyses. The table below shows the size of 
each cluster, which is worth reviewing to check for very small or very large 
clusters that might warrant revisiting your choice of k.
```{r assign-clusters}
# cutree() cuts the dendrogram at k clusters and returns a membership vector
# (one value per student, indicating which cluster they belong to)
cluster_k12 <- cutree(hc, k = 12)

# Add cluster membership to the combined_wide table
combined_wide <- combined_wide %>%
  mutate(cluster = cluster_k12)

# Quick look at cluster sizes
table(cluster_k12) %>%
  as.data.frame() %>%
  rename(Cluster = cluster_k12, N = Freq) %>%
  mutate(Pct = percent(N / sum(N), accuracy = 0.1)) %>%
  arrange(Cluster) %>%
  kable(caption = "Cluster sizes (k = 12)") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

---

# 5. Visualizing clusters

## 5.1 State distribution plots by cluster

The best way to understand what each cluster represents is to plot the state 
distribution for each cluster separately. Looking at these plots will help you
describe the school-to-work pathway(s) captued by each.

**Example Interpretation (Remember Data Are Synthetic)**

-Cluster 1: Students who are primarily in CTC Only and CTC + Work in the early 
quarters, transitioning to Work Only by the middle of the observation window and 
staying there through the end. Likely certificate or associate's degree 
completers who move into stable employment.

-Cluster 2: Students who start with a mix of CTC and work states in the early 
quarters but transition relatively quickly into Work Only, where they remain for 
most of the observation window. Likely short-term CTC enrollees who enter the 
workforce early.

-Cluster 3: Students who are predominantly in 4-Year Only in the early quarters, 
tapering off around Quarter 9 to 13 and transitioning into Neither and some Work 
Only. Likely 4-year enrollees who stop out without completing.

-Cluster 4: Students with a brief period of CTC enrollment in the very early 
quarters followed by a long stretch of Neither. Likely short-term CTC enrollees 
who do not connect to the labor market.

-Cluster 5: Students who are almost entirely in Work Only across all 28 quarters 
with very little variation. Likely students who went directly into the workforce 
after high school and stayed employed throughout.

-Cluster 6: Students with 4-Year Only and 4-Year + Work states in the early 
quarters, transitioning into Work Only around Quarter 13 to 17. Likely  
bachelor's degree completers who move into employment after graduation.

-Cluster 7: Students with a longer stretch of 4-Year Only and 4-Year + Work 
states extending further into the observation window before transitioning to 
Work Only. Likely extended 4-year enrollees or transfer students completing 
bachelor's degrees on a longer timeline.

-Cluster 8: Students with a mix of 4-Year and CTC states in the early quarters, 
with some work, transitioning to Work Only over time. Likely transfer pathway 
students moving between institution types.

-Cluster 9: Students who spend most of the observation window in Neither, with 
very little enrollment or employment. Likely students who are persistently 
disconnected from both school and work.

-Cluster 10: Students who are almost entirely in Work Only from the very first 
quarter, with minimal enrollment. Likely students who went directly into the 
workforce after high school without any postsecondary enrollment.

-Cluster 11: Students with CTC + Work and CTC Only states in the early quarters 
transitioning to Work Only relatively quickly. Likely short-term CTC enrollees 
with concurrent employment who complete a credential and move into the workforce.

-Cluster 12: Students with a longer period of CTC and 4-Year enrollment states, 
including concurrent work, before transitioning to Work Only. Likely transfer 
students or extended CTC enrollees who eventually complete and enter the 
workforce.

```{r cluster-state-dist, fig.height=9}
# seqdplot with group = cluster plots one state distribution panel per cluster
seqdplot(
  seqdata,
  group      = paste0("Cluster ", cluster_k12),
  title      = "State Distribution by Cluster",
  withlegend = "right",
  border     = NA,
  ylab       = "Proportion"
)
```

## 5.2 Index plots by cluster
The plots below also show sequences by cluster, this time with each individual
student's sequence plotted.
```{r cluster-index, fig.height=9}
# Sequence index plots by cluster — shows individual sequences within each cluster
seqiplot(
  seqdata,
  group      = paste0("Cluster ", cluster_k12),
  title      = "Sequence Index Plots by Cluster",
  sortv      = "from.start",
  withlegend = "right",
  border     = NA,
  idxs       = 0
)
```

---

# 6 Labeling clusters as **PATHWAYS**

Looking at the state distribution and index plots, we can assign descriptive 
labels to each cluster/pathway. This is a judgment call and the labels should be:

- **Accurate:** reflecting the dominant sequence pattern
- **Concise:** usable in a table or figure
- **Meaningful to a policy audience:** avoid jargon


You may choose a different cluster solution or choose to highlight different
aspects of each cluster, so your pathway names may be different.

# 6.1 Assign Pathway Labels
```{r label-pathways}
# You can edit the code below with a different set of descriptive pathway labels 

pathway_labels <- c(
  "1"  = "Pathway 1: CTC to Workforce",
  "2"  = "Pathway 2: Brief CTC, Early Workforce Entry",
  "3"  = "Pathway 3: 4-Year Enrollment, No Completion",
  "4"  = "Pathway 4: Brief CTC, Limited Workforce Connection",
  "5"  = "Pathway 5: Direct and Sustained Workforce Entry",
  "6"  = "Pathway 6: 4-Year Completion, Workforce Entry",
  "7"  = "Pathway 7: Extended 4-Year, Workforce Entry",
  "8"  = "Pathway 8: Mixed Enrollment, Workforce Entry",
  "9"  = "Pathway 9: Limited Enrollment and Employment",
  "10" = "Pathway 10: Direct Workforce Entry",
  "11" = "Pathway 11: CTC with Work, Workforce Entry",
  "12" = "Pathway 12: Extended Enrollment, Workforce Entry"
)
# Apply labels to the data
combined_wide <- combined_wide %>%
  mutate(pathway_labels = pathway_labels[as.character(cluster)]) %>% 
  mutate(pathway = cluster)
```


## 6.2 Merge pathway labels onto the analysis flat file

```{r merge-pathways}
# Add pathway membership to the analysis flat file
analysis_with_pathways <- analysis_flat %>%
  left_join(
    combined_wide %>% select(student_id, pathway, pathway_labels),
    by = "student_id"
  )
```

## 6.3 Pathway size summary

```{r pathway-size}
analysis_with_pathways %>%
  count(pathway, pathway_labels) %>%
  mutate(pct = percent(n / sum(n), accuracy = 0.1)) %>%
  arrange(pathway) %>%
  kable(col.names = c("Pathway #", "Label", "N", "% of cohort"),
        caption   = "Pathway sizes") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```
---

# 7. Characterizing pathways by student characteristics and outcomes
Once students have been assigned to pathway groups, the next step is to examine 
who follows each pathway and what outcomes they achieve. This section explores 
how pathway membership varies by student demographic and academic 
characteristics and how labor market outcomes differ across pathway groups. 
These analyses help answer two core research questions: are certain groups of 
students more likely to follow particular pathways, and are certain pathways 
associated with better labor market outcomes? Together, these analyses form the 
substantive heart of the pathway research and are the basis for the policy 
conclusions that can be drawn from this work.

## 7.1 Labor market outcomes by pathway
This table summarizes labor market outcomes by pathway group, showing the number
of students in each pathway, the share who were ever employed during the 
observation window, the share who were stably employed in the final year, and 
median annual wages among stably employed students. 

**Example Interpretation (Remember Data Are Synthetic)**
Pathways involving 4-year enrollment (Pathways 3, 6, and 7) tend to show higher 
median wages among stably employed students, consistent with the earnings 
premium typically associated with bachelor's degree attainment. Pathways with 
limited enrollment and employment connections (Pathways 4 and 9) show lower 
rates of stable employment, reflecting the challenges faced by students who do 
not maintain strong connections to either school or the workforce.

**Note:** These figures are based on fully synthetic data generated for teaching 
and methods demonstration purposes only. The outcome values, including 
employment rates, stable employment rates, and wage figures, are illustrative 
and do not necessarily reflect real labor market conditions in Washington State 
or any other state. Users should expect their results to look meaningfully 
different when working with real administrative data.

```{r outcomes-by-pathway}
analysis_with_pathways %>%
  group_by(pathway, pathway_labels) %>%
  summarise(
    N                 = n(),
    pct_ever_emp      = mean(ever_employed,   na.rm = TRUE),
    pct_stably_emp    = mean(stably_employed, na.rm = TRUE),
    median_wages      = median(annual_wages_2023[stably_employed == 1],
                               na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(pathway) %>%
  mutate(
    `Ever employed`   = percent(pct_ever_emp,   accuracy = 0.1),
    `Stably employed` = percent(pct_stably_emp, accuracy = 0.1),
    `Median wages`    = dollar(median_wages)
  ) %>%
  select(pathway, pathway_labels, N, `Ever employed`, `Stably employed`,
         `Median wages`) %>%
  kable(col.names = c("pathway #", "Label", "N", "Ever employed",
                       "Stably employed", "Median wages (stably emp.)"),
        caption   = "Labor market outcomes by pathway") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE)
```

## 7.2 Wage distributions by pathway (dot plot)
This chart shows the distribution of annual earnings among stably employed 
students by pathway group, sorted from highest to lowest median wages. The dot 
represents the median and the bar spans the 25th to 75th percentile, giving a 
sense of both the typical earnings level and the spread within each pathway.

**Example Interpretation (Remember Data Are Synthetic)**
Pathways involving 4-year enrollment show the highest median wages, with Pathway 
3 (4-Year Enrollment, No Completion) and Pathway 6 (4-Year Completion, Workforce 
Entry) both showing medians above $70,000. Notably, Pathway 3 students who did 
not complete a degree still show relatively high earnings, which may reflect the 
labor market value of some college coursework or the characteristics of students 
who enrolled in 4-year institutions in the first place.

CTC and workforce-focused pathways cluster in the $49,000 to $61,000 range, with 
relatively similar median wages across most groups. Pathway 7 (Extended 4-Year, 
Workforce Entry) stands out as an exception, showing the lowest median wages 
despite involving 4-year enrollment, which may reflect lower credential 
completion rates in that group.

The overlapping interquartile ranges across most pathways are a reminder that 
there is substantial variation within each group, and that pathway membership 
alone does not determine individual outcomes.

**Note:** These figures are based on synthetic data and do not reflect real 
labor market conditions.

```{r earnings-dotplot, fig.height=6}
wages_summary <- analysis_with_pathways %>%
  filter(stably_employed == 1) %>%
  group_by(pathway, pathway_labels) %>%
  summarise(
    p25    = quantile(annual_wages_2023, 0.25, na.rm = TRUE),
    median = median(annual_wages_2023, na.rm = TRUE),
    p75    = quantile(annual_wages_2023, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(pathway_labels = fct_reorder(pathway_labels, median))

ggplot(wages_summary, aes(y = pathway_labels)) +
  geom_segment(aes(x = p25, xend = p75, yend = pathway_labels),
               color = "#AACCEE", linewidth = 4, alpha = 0.7) +
  geom_point(aes(x = median), size = 4, color = "#2E86AB") +
  scale_x_continuous(labels = dollar_format(scale = 1/1000, suffix = "K"),
                     limits = c(0, NA)) +
  labs(
    title    = "Annual Earnings Distribution by pathway",
    subtitle = "Among stably employed students. Bar = 25th-75th percentile; dot = median",
    x        = "Annual wages",
    y        = NULL
  ) +
  theme_minimal(base_size = 13)
```

## 7.3 Percent FRPL by Pathway
This chart shows the share of students in each pathway who received free or 
reduced price lunch (FRPL) in high school, which can be an indicator of economic 
disadvantage. 

**Example Interpretation (Remember Data Are Synthetic)**
FRPL rates vary meaningfully across pathways, ranging from a low 
of 31.3% in Pathway 3 (4-Year Enrollment, No Completion) to a high of 49.1% in 
Pathway 5 (Direct and Sustained Workforce Entry). 

Pathways with lower FRPL rates tend to be those involving 4-year enrollment, 
particularly Pathways 3 and 6, suggesting that economically advantaged students 
are more likely to pursue and persist in 4-year college pathways. Conversely, 
pathways with higher FRPL rates tend to be those with limited postsecondary 
enrollment or direct workforce entry, such as Pathways 5, 4, and 9, which could be
reflecting patterns of economic barriers to postsecondary access.

Pathway 9 (Limited Enrollment and Employment) stands out with the highest FRPL 
rate at 45.8% while also showing the lowest stable employment rate in the 
outcomes table, suggesting that economically disadvantaged students are 
disproportionately represented in the pathway with the most challenging labor 
market outcomes.

**Note:** These figures are based on synthetic data and do not necessarily 
reflect real patterns in Washington State or any other state.

```{r frpl-by-pathway, fig.height=5}
# FRPL rate by pathway
analysis_with_pathways %>%
  group_by(pathway, pathway_labels) %>%
  summarise(frpl_rate = mean(frpl.x, na.rm = TRUE), .groups = "drop") %>%
  mutate(pathway_label = fct_reorder(pathway_labels, frpl_rate)) %>%
  ggplot(aes(x = frpl_rate, y = pathway_labels)) +
  geom_col(fill = "#E84855", alpha = 0.80) +
  geom_text(aes(label = percent(frpl_rate, accuracy = 0.1)),
            hjust = -0.1, size = 3.5) +
  scale_x_continuous(labels = percent_format(), limits = c(0, 0.85)) +
  labs(
    title = "FRPL Rate by pathway",
    x     = "% of pathway receiving free/reduced price lunch",
    y     = NULL
  ) +
  theme_minimal(base_size = 13)
```

# 7.4. Examining differences in pathway participation

This second graph below provides an alternative approach to visualizing 
differences in pathway participation by FRPL status. This chart shows pathway 
participation rates separately for FRPL and non-FRPL students, where each bar 
represents the share of that group following a given pathway. This framing 
highlights differences in how the two groups are distributed across pathways 
rather than just the composition of each pathway.

**Example Interpretation (Remember Data Are Synthetic)**
FRPL students show higher participation in workforce-focused pathways with 
limited postsecondary engagement, particularly Pathway 5 (Direct and Sustained 
Workforce Entry, 24.0% vs. 18.9%), Pathway 10 (Direct Workforce Entry, 12.5% vs. 
10.8%), and Pathway 4 (Brief CTC, Limited Workforce Connection, 10.6% vs. 9.3%). 
Non-FRPL students show higher participation in pathways involving sustained 
postsecondary enrollment, particularly Pathway 6 (4-Year Completion, Workforce 
Entry, 9.0% vs. 6.7%) and Pathway 3 (4-Year Enrollment, No Completion, 8.7% vs. 
5.2%). Differences across most CTC pathways are smaller.

**Note:** These figures are based on synthetic data and do not necessarily 
reflect real patterns in Washington State or any other state.

```{r pathway-by-frpl, fig.height=8, fig.width=12}
# Identify "high-earnings" pathways as those with median wages above the
# overall median. In a real analysis, you would use labeled pathways here.
overall_median_wage <- median(
  analysis_with_pathways$annual_wages_2023[
    analysis_with_pathways$stably_employed == 1
  ], na.rm = TRUE
)

analysis_with_pathways %>%
  group_by(frpl = factor(frpl.x, levels = c(0, 1),
                          labels = c("Non-FRPL", "FRPL")),
           pathway_labels) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(frpl) %>%
  mutate(pct = n / sum(n)) %>%
  ungroup() %>%
  mutate(pathway_label = fct_reorder(pathway_labels, pct)) %>%
  ggplot(aes(x = pct, y = pathway_labels, fill = frpl)) +
  geom_col(position = "dodge", alpha = 0.85) +
  geom_text(aes(label = percent(pct, accuracy = 0.1)),
            hjust = -0.1, size = 3.5, position = position_dodge(width = 0.9)) +
  scale_x_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("Non-FRPL" = "#2E86AB", "FRPL" = "#E84855")) +
  labs(
    title = "Pathway Participation by FRPL Status",
    x     = "Share of group in this pathway",
    y     = NULL,
    fill  = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```



## 7.5 HS GPA distributions by pathway

High school GPA is a strong predictor of postsecondary pathway. This ridge plot 
shows the distribution of high school GPA for students in each pathway group, 
sorted from highest to lowest median GPA. Each ridge represents the full 
distribution of GPAs within a pathway, making it possible to see not just the 
typical GPA but also the spread and shape of the distribution.

**Example Interpretation (Remember Data Are Synthetic)**
Pathways involving 4-year college enrollment are concentrated at the top of the 
chart, with Pathway 6 (4-Year Completion, Workforce Entry) and Pathway 3 (4-Year 
Enrollment, No Completion) showing the highest GPA distributions, peaking around 
3.5. CTC pathways cluster in the middle, with distributions centered around 2.5 
to 3.0. Pathways with limited or no postsecondary enrollment, particularly 
Pathway 5 (Direct and Sustained Workforce Entry) and Pathway 11 (CTC with Work, 
Workforce Entry), show the lowest and widest GPA distributions, with more 
students in the 1.5 to 2.5 range.

The overlap between pathway distributions is worth noting. There are students 
with high GPAs in every pathway, including workforce-only pathways, and 
students with lower GPAs in the 4-year pathways. This suggests that while prior 
academic preparation is strongly associated with pathway participation, it does 
not fully determine it, and that factors beyond GPA also play important roles in
shaping which pathway a student follows.

**Note:** These figures are based on synthetic data and do not necessarily 
reflect real patterns in Washington State or any other state.
```{r gpa-ridgeplot, fig.height=7}
hs_records <- read_csv(file.path(data_dir, "hs_records.csv"))

analysis_with_pathways %>%
  left_join(hs_records %>% select(student_id, hs_gpa), by = "student_id") %>%
  filter(is.na(hs_gpa.x)==0) %>%
  mutate(pathway_labels = fct_reorder(pathway_labels, hs_gpa.x, median)) %>%
  ggplot(aes(x = hs_gpa.x, y = pathway_labels, fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 1.5, rel_min_height = 0.01) +
  scale_fill_gradient(low = "#FFE0B2", high = "#1B6CA8") +
  labs(
    title = "High School GPA Distribution by pathway",
    x     = "High school GPA",
    y     = NULL,
    fill  = "GPA"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none")
```


---

# 8. Saving final results
This final step saves the analysis dataset with pathway membership appended as a 
new column. The resulting file, final_results_with_pathways.csv, contains one 
row per student and includes all demographic characteristics, high school 
records, credential attainment, labor market outcomes, and each student's 
assigned pathway group. This file can be used for any subsequent analyses 
outside of this module, such as regression models, subgroup comparisons, or data 
visualizations, without needing to re-run the clustering steps.
```{r save-results}
# Save the final pathway membership merged with the analysis flat file
final_results <- analysis_with_pathways

write_csv(final_results, file.path(data_dir, "final_results_with_pathways.csv"))

cat("Saved final_results_with_pathways.csv to", data_dir, "\n")
cat("Rows:", nrow(final_results), "\n")
cat("Columns:", ncol(final_results), "\n")
```

---

# 9 Summary and key takeaways

## What we did

This three-module series demonstrated a complete sequence + cluster analysis 
workflow applied to synthetic state longitudinal data:

1. **Module 1:** Loaded and explored the raw longitudinal data tables; 
   produced descriptive statistics on demographics, enrollment, credentials, 
   and labor market outcomes.

2. **Module 2:** Defined sequences using TraMineR; visualized trajectory 
   patterns; computed Optimal Matching pairwise distances.

3. **Module 3:** Applied Ward hierarchical clustering to the distance matrix; 
   selected k using quality criteria; labeled pathways; validated against known 
   archetypes; characterized pathways by demographics and outcomes.

## Reminder of Key methodological decisions

Every step in this workflow involves choices that affect results. Here are some 
of the most consequential:

- State definition: We used 6 combined enrollment and employment states for the 
module code, though the larger Sequences to Success in WA State project ran 
several version of the states, including combined enrollment and employment. We
ultimately landed on a set of states that embedded credential attainment and 
prioritized enrollment over employment, since this made the most sense for our 
team's research goals. Other alternatives include using enrollment status only, 
or creating finer-grained employment states that distinguish part-time from 
full-time work or earnings levels.

- Distance method: We used Optimal Matching (OM) with TRATE substitution costs, 
which weights substitutions by the observed frequency of transitions between 
states in the data. Alternatives include DHD, LCS, and OMslen, each of which 
emphasizes different aspects of sequence similarity (see Section 6.3 for a 
fuller discussion).

- Indel cost: We set the indel cost to 1.5, which penalizes insertions and 
deletions more than the default and tends to produce distances that are more 
sensitive to the timing of transitions. Alternatives of 1.0 or 2.0 are worth 
testing as a sensitivity check to confirm that your substantive conclusions do 
not change meaningfully with different cost settings.

- Clustering algorithm: We used Ward linkage hierarchical clustering, which 
tends to produce compact, evenly sized clusters and is the most common approach 
for sequence analysis with OM distances. Alternatives include complete linkage, 
which tends to produce unequal cluster sizes, and k-medoids, which can be more 
robust but is slower on large datasets.

- Number of clusters: We chose 12 clusters based on a combination of cluster 
quality statistics and substantive knowledge of the pathways we expected to find. 
There is no single correct answer, and the choice should be guided by both 
statistical fit and interpretability for your research audience.

When reporting results, be transparent about these choices and, where possible, 
show that your conclusions are robust to alternatives.


**You have now completed the full sequence and cluster analysis workflow.** 
The following file has been saved to your longitudinal_toy_data/ folder:

- final_results_with_clusters.csv — the complete analysis dataset with pathway 
membership appended, ready for any subsequent analyses such as regression models, 
subgroup comparisons, or data visualizations.

This is the final module in the series. If you would like to adapt this workflow 
to your own state longitudinal data, return to 00_building_your_analysis_file.Rmd 
for guidance on transforming raw enrollment and wage records into the 
sequence-ready format used across these modules.

---

# Session information

The code in this module was developed and tested using the package versions 
listed below. If you encounter unexpected errors, checking your package 
versions against these is a good first troubleshooting step. You can update 
packages by running `update.packages()` in your RStudio console.

```{r session-info}
sessionInfo()
```

**Development Session Info**
R version 4.5.2 (2025-10-31)
Platform: x86_64-pc-linux-gnu
Running under: Red Hat Enterprise Linux 9.8 (Plow)

Matrix products: default
BLAS/LAPACK: FlexiBLAS OPENBLAS-OPENMP;  LAPACK version 3.9.0

locale:
LC_CTYPE=C.UTF-8       LC_NUMERIC=C           LC_TIME=C.UTF-8        LC_COLLATE=C.UTF-8     LC_MONETARY=C.UTF-8   
LC_MESSAGES=C.UTF-8    LC_PAPER=C.UTF-8       LC_NAME=C              LC_ADDRESS=C           LC_TELEPHONE=C        
LC_MEASUREMENT=C.UTF-8 LC_IDENTIFICATION=C   

time zone: America/New_York
tzcode source: system (glibc)

attached base packages:
stats     graphics  grDevices utils     datasets  methods   base     

other attached packages:
ggridges_0.5.7        WeightedCluster_1.8-1 cluster_2.1.8.1       RColorBrewer_1.1-3    TraMineR_2.2-12      
kableExtra_1.4.0      knitr_1.50            scales_1.4.0          lubridate_1.9.4       forcats_1.0.1        
stringr_1.6.0         dplyr_1.1.4           purrr_1.2.0           readr_2.1.6           tidyr_1.3.1          
tibble_3.3.0          ggplot2_4.0.1         tidyverse_2.0.0      

loaded via a namespace (and not attached):
tidyselect_1.2.1    viridisLite_0.4.2   farver_2.1.2        S7_0.2.1            fastmap_1.2.0       digest_0.6.38      
timechange_0.3.0    lifecycle_1.0.4     magrittr_2.0.4      compiler_4.5.2      rlang_1.1.6         sass_0.4.10        
tools_4.5.2         yaml_2.3.10         labeling_0.4.3      bit_4.6.0           xml2_1.4.1          rsconnect_1.6.1    
withr_3.0.2         nnet_7.3-20         grid_4.5.2          colorspace_2.1-2    future_1.67.0       progressr_0.18.0   
fastcluster_1.3.0   globals_0.18.0      iterators_1.0.14    MASS_7.3-65         cli_3.6.5           rmarkdown_2.30     
vegan_2.7-2         crayon_1.5.3        generics_0.1.4      rstudioapi_0.17.1   future.apply_1.20.0 tzdb_0.5.0         
cachem_1.1.0        splines_4.5.2       parallel_4.5.2      vctrs_0.6.5         vegclust_2.0.3      boot_1.3-32        
Matrix_1.7-4        jsonlite_2.0.0      hms_1.1.4           bit64_4.6.0-1       listenv_0.10.0      systemfonts_1.3.1  
foreach_1.5.2       jquerylib_0.1.4     glue_1.8.0          parallelly_1.45.1   codetools_0.2-20    stringi_1.8.7      
gtable_0.3.6        doFuture_1.1.2      pillar_1.11.1       htmltools_0.5.8.1   R6_2.6.1            textshaping_1.0.4  
vroom_1.6.6         evaluate_1.0.5      lattice_0.22-7      bslib_0.9.0         svglite_2.2.2       nlme_3.1-168       
permute_0.9-8       mgcv_1.9-3          xfun_0.54           pkgconfig_2.0.3    

---

*Synthetic data generated for teaching purposes.*
